---
title: "556TermProject_Proposal"
authors: "Callan, Abbie, and Katie - ENV H 556"
format:
  html:
    df_print: "paged"
    fig_caption: yes
    toc: true
    toc_depth: 3
    number_sections: true
    self-contained: true #save images etc. in this file (vs folders)
execute:
  echo: true
  cache: false
  echo.comments: false
  message: false
  warning: false
  
---

This document was rendered on `r format(Sys.time(), '%B %d, %Y')`.
---

## Setting up

For our term project, we aim to assess the temporal variations of traffic-related air pollution (TRAP). Specifically, we will assess how air pollution levels vary across seasons, days of the week, or hours of the day. Our project will use data from the Mobile Monitoring Campaign, as described by [Blanco et al.](https://pubmed.ncbi.nlm.nih.gov/35917479/). 



```{r setup, include=FALSE}

#-----setup-----

# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
   
}

```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}

#-----load libraries pacman-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.
pacman::p_load(knitr, kableExtra, tidyverse, egg, multcomp, modelr,
               broom # tidy() # optional
               )

```

```{r directory.organization.read.data, echo=FALSE, warning=FALSE}

#-----directory organization and read data-----

# specify data path, data will be added to your local working directory, should be the project folder
data_path <- file.path(getwd())

# specify the file name and path
file_name <- "allseasonsR.rds"
file_path <- file.path(data_path, file_name)

# Download the file if it is not already present
if (!file.exists(file_path)) {
    url <- paste("https://faculty.washington.edu/sheppard/envh556/Datasets", 
                 file_name, sep = '/')
    download.file(url = url, destfile = file_path)
}

# Output a warning message if the file cannot be found
if (file.exists(file_path)) {
    snapshot<-readRDS(file_path)
} else warning(paste("Can't find", file_name, "!"))

# remove temporary variables
rm(url, file_name, file_path, data_path)

```

## Introduction

*Introduction to Mobile Monitoring Data set and variables*


*specific aims/hypotheses*





## Methods & Statistical Approach




