---
title: "556TermProject_Proposal"
authors: "Callan, Abbie, and Katie - ENV H 556"
format:
  html:
    df_print: "paged"
    fig_caption: yes
    toc: true
    toc_depth: 3
    number_sections: true
    self-contained: true #save images etc. in this file (vs folders)
execute:
  echo: true
  cache: false
  echo.comments: false
  message: false
  warning: false
  
---

This document was rendered on `r format(Sys.time(), '%B %d, %Y')`.
---

## Setting up

For our term project, we aim to assess the temporal variations of traffic-related air pollution (TRAP). Specifically, we will assess how air pollution levels vary across seasons, days of the week, or hours of the day. Our project will use data from the Mobile Monitoring Campaign, as described by [Blanco et al.](https://pubmed.ncbi.nlm.nih.gov/35917479/). 



```{r setup, include=FALSE}

#-----setup-----

# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
   
}

```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}

#-----load libraries pacman-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.
pacman::p_load(knitr, kableExtra, tidyverse, egg, multcomp, modelr,
               broom, EnvStats )

```

```{r directory.organization.read.data, echo=FALSE, warning=FALSE}

#-----directory organization and read data-----

# specify data path, data will be added to your local working directory, should be the project folder
data_path <- file.path(getwd())

# specify the file name and path
file_name <- "allseasonsR.rds"
file_path <- file.path(data_path, file_name)

# Download the file if it is not already present
if (!file.exists(file_path)) {
    url <- paste("https://faculty.washington.edu/sheppard/envh556/Datasets", 
                 file_name, sep = '/')
    download.file(url = url, destfile = file_path)
}

# Output a warning message if the file cannot be found
if (file.exists(file_path)) {
    snapshot<-readRDS(file_path)
} else warning(paste("Can't find", file_name, "!"))

# remove temporary variables
rm(url, file_name, file_path, data_path)

```


```{r}
#-----read data from a website--------

  
```



## Introduction

*Introduction to Mobile Monitoring Data set and variables*
  The Mobile Monitoring dat set comes from a study of air pollution and its association with ageing. The study collected air samples
  from around the greater Seattle area (Washington, United States). A car, equipped with air quality monitoring and sampling
  equipment, drove along nine routes, stopping and taking ~ 309 samples across all nine routes. Each sample is a two-minute sample
  of air quality, including pollutant concentrations for nitrogen dioxide (NO2), PM2.5, particle number concentration (PNC), carbon
  dioxide (CO2), and black carbon (BC). This air sample data was used to associate with ageing outcomes in study participants. Our
  report will not look at any health outcomes, and will focus on characterizing the temporal patterns in the air pollution data. 
  Over the course of a year, each site along the nine driving routes was visited about 25 times [Blanco et al., 2022].

*specific aims/hypotheses*

  The sampling design for the Mobile Monitoring data set included variability in the day time, week, and season that samples
  were taken for a particular site. This means that each sampling location has sampling data from different times throughout 
  the day, different samples tghroiughout the week, and multiples throughout the year. Our report will utilize the multiple 
  layers of temporal variation in sample collection to assess changes in air quality across a week and across a year. 
  
  We hypothesize that ___________________________________.



## Methods & Statistical Approach






## Exploratory/ Introductory Analysis

Table 1.0

```{r, echo = FALSE, warning = FALSE, message = FALSE}

kable(annual%>%
  group_by(variable) %>%
  summarise(
    n = length(value),  #every value is 309, can omit and summarize in text. 
    nmiss = sum(is.na(value)), #no missing, can omit and summarize in text.
    GM = geoMean(value, na.rm = TRUE),
    GSD = geoSD(value, na.rm = TRUE),
    AM = mean(value, na.rm = TRUE),
    ASD = sd(value, na.rm = TRUE)),
  digit = 1)

```
```{r}
#---Q-Q plot---#

#using a qq plot to assess normality

#plotting first on a native scale
ggplot(annual, aes(sample = value)) +
  stat_qq() + 
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot of TRAP",
       x = "Theoretical Quantiles from a Std Normal Distribution",
       y = "Concentration - native scale"
       ) +
  facet_wrap(~variable, scales = "free")

#Co2 and PM2.5 look close to normal on the native scale, although there are some slight deviations at the tails. 
#other trap have significant deviation, mostly in the upper quantiles

#creating log transformed concentration variable. 
annual <- annual %>% mutate(logvalue = log(value), .after = value)

#qq plots using log transformed values
ggplot(annual, aes(sample = logvalue)) +
  stat_qq() + 
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot of TRAP",
       x = "Theoretical Quantiles from a Std LogNormal Distribution",
       y = "log(concentration)"
       ) +
  facet_wrap(~variable, scales = "free")

#CO2 nad PM2.5- also looks ok lognormal
#ma200 - looks much better lognormal
#No2, ns, pmdisc, pnc - still doesn't look normal, although the deviation is less severe than before transformation

```


