---
title: "CA NOx Term Project"
authors: "Callan, Abbie, and Katie - ENV H 556"
format:
  html:
    df_print: "paged"
    fig_caption: yes
    toc: true
    toc_depth: 3
    number_sections: true
    self-contained: true #save images etc. in this file (vs folders)
execute:
  echo: false
  cache: false
  echo.comments: false
  message: false
  warning: false
  
---

This document was rendered on `r format(Sys.time(), '%B %d, %Y')`.
---

# Introduction and Purpose

Nitrogen oxides (NOx) are gaseous compounds commonly found in air pollution. Although NOx are generated through both natural and anthropogenic sources, human activities account for the majority of ambient NOx concentrations [(EPA 2024;)](https://www.epa.gov/no2-pollution/basic-information-about-no2) [Zhang et al. 2003;](https://www.pnas.org/doi/10.1073/pnas.252763799) [(ATSDR 2002;](https://www.atsdr.cdc.gov/toxfaqs/tfacts175.pdf) [(EPA 1999)](https://www3.epa.gov/ttncatc1/dir1/fnoxdoc.pdf). Motor vehicles are a major source of anthropogenic NOx, accounting for approximately half of all emissions related to human activities [(EPA 1999)](https://www3.epa.gov/ttncatc1/dir1/fnoxdoc.pdf). Power plants also contribute a substantial fraction (~20%) of the total estimated annual anthropogenic NOx emissions [(EPA 1999)](https://www3.epa.gov/ttncatc1/dir1/fnoxdoc.pdf).

Traffic-related air pollutants (TRAP), such as NOx, can have diurnal and seasonal trends related to temporal vehicular traffic patterns [(Blanco et al. 2022)](https://www.nature.com/articles/s41370-022-00470-5#MOESM1). Measurements of TRAP must therefore appropriately reflect the relevant temporal period that they are intended to represent. Traditionally, air pollution is measured through stationary area monitors, which provide continuous air concentration data over long time scales from a single location [(CDPHE 2024)](https://cdphe.colorado.gov/air-toxics/monitoring#:~:text=Mobile%20monitoring%20can%20capture%20%E2%80%9Cspatial,wind%20compared%20to%20stationary%20monitoring.). Mobile monitoring is another effective air pollution measurement strategy that can be used to evaluate air concentrations across a wide set of geographic areas. During mobile monitoring, vehicles equipped with air sampling equipment drive through areas and provide real-time measurements of air pollution across a series of stops (CDPHE 2024). Mobile monitoring is an especially advantageous air monitoring strategy due to its spatial flexibility, which can provide data on areas not well-represented by stationary monitoring sites (CDPHE 2024). Despite its spatial advantages, mobile monitoring is more resource intensive than stationary monitoring and therefore provides a more temporally limited set of air monitoring data (CDPHE 2024). 

In this analysis, we use the California NOx dataset described by Blanco et al. (2023) to simulate a series of temporally-restricted mobile monitoring campaigns and compare their prediction performances against that of a long-term, temporally-balanced "gold standard" dataset representing what might be achieved with a stationary air monitor. We aimed to assess how well NOx predictions from short-term mobile monitoring campaigns aligned with the "gold standard" dataset and whether the inclusion or omission of specific temporal windows impacted prediction performance. In support of this aim, we compared the performance of a "gold standard" temporally-balanced statewide dataset with a (1) short-term, temporally-balanced statewide dataset; and (2) short-term, seasonally-restricted statewide datasets. Additionally, we repeated this analysis when all data were instead restricted to Los Angeles county to assess how the comparisons differed in the presence of reduced geospatial variability. Ultimately, these results can be used to inform future TRAP monitoring campaigns through signaling the importance of temporally-balanced sampling designs and underscoring which key temporal windows must be included in mobile monitoring sampling events. 

# Methods 

**Dataset Description**

The California NOx dataset, first described by [Blanco et al. (2023)](https://www.nature.com/articles/s41370-022-00470-5#MOESM1), includes NOx concentration measurements collected from 69 California Air Quality System (AQS) sites in 2016. Measurements were collected every hour, enabling the evaluation of temporal trends in NOx concentrations across both short- and long-term timescales. Land-use, roadway proximity, and population density covariates were additionally available in the dataset.  

**Overview of Statistical Approach** 

We developed a general model for NOx predictions to be used across all analyses. We evaluated out-of-sample NOx prediction performances using the general model for our "gold standard" temporally-balanced dataset, our short-term temporally balanced dataset, and our short-term temporally restricted datasets. Predictions from each temporally subset dataset were evaluated against observations from the gold standard data as well as observations from each subset to control for bias in the observations as a result of sampling from temporally restricted data. Details regarding the creation and characteristics of these datasets are provided in subsequent sections. 

For all analyses, NOx concentrations were modeled on the natural log scale based on the methods described by Blanco et al. (2023) and based on our visual inspection of the variable distributions (See Appendix for Distributions). 

*Sensor Selection*

Sensors from the California NOx dataset were selected for inclusion in our analysis according to the criteria described by Blanco et al. (2023). These criteria included: (1) limited missingness, such that the sensors included had annual data that was at least 66% complete; (2) limited data gaps, such that data gaps for a particular sensor were less than or equal to 45 days long; and (3) monitors must have positive readings (> 0 ppb) at least 60% of the time. Negative concentration readings were left in the data set, as we interpreted them as true reading close to 0 that reads as negative due to noise or "classical error" in the instrument readings. We restricted the percentage of negative readings in each sensor to ensure that the sensors included had adequate variability in their readings and that annual average concentrations would be positive. 

*Model Selection*

Covariates were selected for inclusion into our general model for NOx predictions based on their scientific relevance to ambient NOx concentration prediction models, as described in peer-reviewed scientific literature. Specifically, we chose to include the covariates utilized by [Mercer et al. (2011)](https://pmc.ncbi.nlm.nih.gov/articles/PMC3146303/pdf/nihms299256.pdf), in the Snapshot campaign that measured ambient NOx concentrations in Los Angeles across seasons and built prediction models using land-use and geographic covariates. Mercer et al. (2011) selected their prediction model covariates from 65 possible covariates related to population density, land-use intensity, open space land, distance to the coast, distance to industrial sources, distances to various roadways, and lengths of roadways within a given buffer zone. From these potential covariates, Mercer et al. (2011) created a "common model" that had adequate performance in predicting NOx concentrations across all seasons. The covariates included in the Mercer et al. (2011) common model included: distance to commercial pollutant sources, distance to the coast, distance to A1 roadways, population size within 5000 meters, land use intensity within 3000 square kilometers, length of A1 roads within 50 meters, and length of A2 and A3 roads within 400 meters.

For our analyses, we used forward stepwise regression to select covariates for inclusion into our model from those most similar to the seven used by Mercer et al. (2011). The Mercer et al. (2011) covariates were deemed appropriate to use based on the similarities in geographic region, pollutant of interest, land-use covariates, and modeling goals between the present analysis and that described by Mercer et al. (2011). These covariates included: meters to closest commercial and services area, distance to the closest coastline, population density within 5,000 meters, distance to nearest A1 roadway, length of A1 roadway within 500 meters, length of A2 roadway within 1500 meters, length of A3 roadway within 400 meters, and the proportion of mixed urban of built-up land within a 1500 square kilometer buffer. Note that, in some cases, distances and buffers varied between those used by Mercer et al. (2011) and those used by Blanco et al. (2023); in these cases, the closest available covariate distance or buffer was selected. For land-use intensity, the proportion of mixed urban or built-up land within 1500 square kilometers was used in the present analysis, as it was deemed to be likely the most similar to the Mercer et al. (2011) land use intensity covariate. 

*Gold Standard Sampling Dataset*

To make our temporally-balanced "gold standard" sampling dataset, we included all measurements from monitors that met our pre-defined inclusion criteria. The gold standard dataset therefore reflected an ideal sampling mechanism, representing sample collection at every hour on every day in all seasons during 2016. As noted by Blanco et al. (2023), the gold standard dataset is reasonably representative of the measurements one might obtain from annual stationary air monitoring. All measurements from each monitor were aggregated into an annual average NOx concentration in ppb for each monitor. 

We predicted the annual average NOx concentration using our gold standard sampling dataset with 10-fold cross-validation. We computed the MSE-based $R^2$ and RMSE characterizing the model fit and accuracy within this dataset. 

*Short-Term Sampling Dataset*

We created short-term sampling datasets intended to represent mobile monitoring campaigns within short-term sampling periods. Our short-term sampling datasets included temporally balanced and temporally restricted subsets of the data. The temporally balanced subset included randomly selected measurements from all measurements in the gold standard dataset. For our temporally-restricted datasets, we stratified our dataset by season and restricted measurements for each monitoring campaign to a single season; this was intended to simulate a mobile monitoring campaign collecting all data on an expedited timeframe with little seasonal variability in the collected samples. 

For each short-term sampling period, we filtered the dataset to include only samples from the specified timeframe, randomly sampled 28 measurements from each monitor, and repeated this random selection 30 times. 28 measurements were selected based on findings in Blanco et al. (2023), which reported that 28 measurements could accurately estimate the annual average of a site (less than or equal to 25% error) and reasonably represented the sampling design of a single mobile monitoring campaigns, which collect repeated samples from a limited number of locations. The random sampling process was repeated 30 total times to provide additional robustness through simulating multiple mobile monitoring campaigns, as described by Blanco et al. (2023). For the short-term campaigns, the 28 sampled measurements from each monitor were aggregated to compute an annual average NOx concentration for that campaign. This "annual average" different from the true annual of average (annual average of the gold standard data), because it was based on only on 28 samples per monitors rather than all available data. 

For each short-term sampling dataset, we predicted the expected annual average NOx concentration with 10-fold **do we want to do LOO for all for consistency** cross-validation based on the results from all 30 sampling iterations. 

*Comparison of Sampling Approaches*

For each of the short-term datasets, we computed the expected value of MSE-based $R^2$ and RMSE based on all 30 sampling iterations comparing the cross-validated predictions from that sampling campaign to the observations from the same campaign.  We additionally computed performance statistics (RMSE and $R^2$) and compared the cross-validated annual average estimates from the temporally-restricted sampling datasets with that estimated using the gold standard sampling dataset. Since we are evaluating the ability of a short-term campaign to predict annual average concentrations (the primary metric of interest for epidemiologic studies), comparing the short-term campaign predictions to the gold standard observations controls for bias in the sampling method itself. 

*Sensitivy Analysis: Los Angeles County*

In our sensitivty analysis, we repeated our analyses with geographically restricted data from a single county to limit the impact of geospatial variability, as NOx concentration measurements likely have strong spatial correlation at the county-level based on shared land-use, policy, and urbanicity within counties. Los Angeles County was selected as the county of interest for our analysis based on its high population density (U.S. Census 2023) and because of its high proportion of sensors included in the underlying California NOx dataset. Because there were fewer sensors in Los Angeles County relative to the statewide dataset, 10-fold cross validation was not possible in this sensitivity analysis. Instead, we used leave-one-out cross validation to validate Los Angeles-specific NOx predictions using the Los Angeles only monitors from the same sampling campaigns used the main analysis. 



# Results

In total, 69 sensors in the state of California met the specified inclusion criteria and were included in our NOx prediction analysis (Figure 1). For our analyses specific to Los Angeles county, 12 sensors meeting the inclusion criteria were included. The distribution of NOx concentrations in our data set were log-normally distributed, as illustrated in Appendix figure 2. 

```{r setup, include=FALSE}

#-----setup-----

# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
   
}

```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}

#-----load libraries pacman-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# **SPH server**: need to install rnaturalearthhires like so on the SPH server
if (!require("rnaturalearthhires")) {
    install.packages("rnaturalearthhires", repos = "https://ropensci.r-universe.dev", type = "source")
} #put this inthe top of the main doc


# Load the other packages, installing as needed.
pacman::p_load(knitr, kableExtra, tidyverse, lubridate, egg, multcomp, modelr, broom, EnvStats, Hmisc,dplyr, tidyr, purrr, ggplot2, stringr, sf, lme4, VCA, gridExtra, ggspatial, maptiles, rnaturalearth, rnaturalearthhires, gstat, prettymapr, patchwork)


```


```{r read.data, echo=FALSE, include=FALSE}
#-----read data from a website--------

# create data directory if it does not exist
dir.create(file.path("Datasets"),
           showWarnings = FALSE,
           recursive = TRUE)
data_path <- file.path("Datasets")
grid.file <- "la_grid_3_5_19.csv"
grid.path <- file.path(data_path, grid.file)

# read in ca nox air pollution concentrations from mobile monitoring
# download it from the web and save locallyif the file does not already exist
# read data; download if necessary
cal_nox <- read_rds(file.path("https://zenodo.org/records/14166411/files/nox_hourly.rda?download=1", 
                                 output_file_path = file.path("Datasets", "nox_hourly.rda"))) %>% rename_with(~ tolower(gsub(".","_", .x, fixed=TRUE)))

#view the data
glimpse(cal_nox)
summary(cal_nox)

length(unique(cal_nox$native_id)) #73 locations

unique(cal_nox$parameter_name) #3 different measurements NO, NO2, NOX


#importing covariate data
                              
ca_covariates <- read_rds(file.path("https://zenodo.org/records/14166411/files/site_covariates.rda?download=1"))

#glimpse(ca_covariates) #too long!

#going to do the attaching later because it creates such a huge df

#importing grid data


# save coordinate systems as variables
  # WGS84 latitude-longitude
latlong_proj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
  # Lambert Conic projection (meters)
lambert_proj <- "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"

if (!file.exists(grid.path)) {
    url <- paste("https://faculty.washington.edu/sheppard/envh556/Datasets", 
                 grid.file, sep = '/')
    download.file(url = url, destfile = grid.path)
}

if (file.exists(grid.path)) {
    la_grid <- read_csv(file = grid.path)
} else warning(paste("Can't find", grid.file, "!"))

# Check initial class of la_grid
class(la_grid)

# Filter out rows with -Inf in D2A1, remove redundant lambert columns, and convert to sf
la_grid <- la_grid %>%
  filter(D2A1 != -Inf) %>%
  dplyr::select(-lambert_x, -lambert_y) %>%
  st_as_sf(coords = c('longitude', 'latitude'), crs = latlong_proj)  

#making the covariates data a sf 

ca_covariates <- st_as_sf(ca_covariates, coords = c("longitude", "latitude", crs = latlong_proj)) 


  


```



```{r create.strata, echo =F, include = F}

#adding variables to use for subsetting

cal_nox <- cal_nox %>%
  mutate(day_time = case_when(
    hour(date) > 4 & hour(date) <= 9 ~ "Morning",
    hour(date) > 9 & hour(date) <= 16 ~ "Midday",
    hour(date) > 16 & hour(date) <= 21 ~ "Evening",
    hour(date) > 21 | hour(date) <= 4 ~ "Night"
  ), .after = hour) %>%
  mutate(season = factor(season, levels = c("Morning", "Midday", "Evening", "Night")))

#I made these cuts off very rough estimates based on data distribution. We probably want to fine tune our scientific rationale for doing so and readjust if warranted. KW

#fixing the season variable because there were a lot of NA

cal_nox <- cal_nox %>%
  mutate(season = case_when(
    week(date) > 12 & week(date) <= 25 ~ "Spring",
    week(date) > 25 & week(date) <= 38 ~ "Summer",
    week(date) > 38 & week(date) <= 50 ~ "Fall",
    week(date) > 50 | week(date) <= 12 ~ "Winter"
  )) %>%
  mutate(season = factor(season, levels = c("Spring", "Summer", "Fall", "Winter")))


#summarizing by Day of the Week
cal_nox<- cal_nox %>%
  mutate(weekday2 = as.character(weekday) %>%
           str_replace(pattern = "TRUE",
                       replacement = "Weekday") %>%
         str_replace(pattern = "FALSE",
                     replacement = "Weekend"), .after = weekday)


```

```{r, add.covars, echo = F, include = F}

#Attaching covariates to the data with all the temporal variables added. Use this if needed for model selection
# combine files
cal_nox_covars <- left_join(cal_nox, ca_covariates, by="native_id")
```



```{r cleaning.data, echo = F, include = F}

#cleaning the data for sampling per the Blanco et al. paper

#create a df that is only NOX

nox_only <- cal_nox %>% filter(parameter_name == "Oxides of nitrogen (NOx)")


length(unique(nox_only$native_id)) #There are 69 monitors here, so we lost a couple when we filtered it to NOX only (there were 73 monitors when NO and NO2 were also included). 

#determining missingness
lapply(nox_only, function(i){ 
   tibble( 
          # sum missing
          n_miss = sum(is.na(i)), 
          
          # percent missing
          perc_miss = round(n_miss/length(i) * 100, 1)
          )
   }) %>% 
   # bind list
   bind_rows(.id = "variable")
# There are no NA values, but not sure if there is a complete record for each monitor

#calculate the total number of expected measurements per monitor

#check the range of the data
min(nox_only$date)
max(nox_only$date)
#The data ranges from "2016-01-01 00:00:00 PST" to "2016-12-31 23:00:00 PST"

#confirm 2016 is a leap year
leap_year(2016) # it is, and there are 366 DOY

#expected number of measurements per monitor: 366 days * 24 measurements/day
meas <- 366 *24 #expect 8784 measurements/monitor if there are no gaps for the year

meas_summary <- nox_only %>% 
  group_by(native_id) %>% 
  summarise(
    count = n(), 
    pct_total = n()/meas*100, 
    pct_pos = sum(sample_measurement > 0)/n()*100)

min(meas_summary$pct_total) #77.8% of the year
max(meas_summary$pct_total) #96.9% of the year

#all of this meets the first criteria in the Blanco paper, which is to have reading >66% of the year

#second criteria from Blanco: data gaps <= 45 days long

# Initialize a dataframe
max_gap <- data.frame(native_id=unique(nox_only$native_id), max_gap=NA)

# Loop over the monitors to calculate the maximum gap in measurements in days
for(i in 1:nrow(max_gap)){
  data.i <- subset(nox_only, native_id == max_gap$native_id[i])
  data.i <- data.i[order(data.i$doy), ]  # Ensure data is sorted by 'doy'
  
  gaps <- diff(data.i$doy)  # Calculate the differences between consecutive days
  max_gap$max_gap[i] <- max(gaps, na.rm = TRUE)  # Store the maximum gap
}

max(max_gap$max_gap) #maximum number of consecutive days w/o measurements is 36

#based on blanco criteria there is no need to eliminate any monitors

#Third Blanco criteria is that the monitor sampled for 40% of the time during the two week period used in "common design" sampling models. Will implement this later if we choose to do a 2 weeks sampling window. 

#Fourth Blanco critera is the monitor is >0 60% of the time. 

min(meas_summary$pct_pos) #70.6%
max(meas_summary$pct_pos) #100%

#no need to eliminate any monitors based on this criteria. 

#No monitors were eliminated based on this cleaning. 

##**Just a thought if we have time - could put a table of monitor characteristics (percenting missigness, etc) in the appendix**##
##*KW will make this is everything else comes together. 

sensor_summary <- left_join( meas_summary, max_gap,  by = "native_id") 
#Abbie put a table of data gaps and sensor criteria in the appendix


```


```{r,random.sampling fxn, echo = F, include = F}
#per Blanco, we will pick 28 samples per monitor. The Gold Standard sample, will be distributed across the whole year. We will then do temporally restricted samples to see how they compare

#create factor to make subsetting in the loop easier for resample
nox_only$native_id_fact <- factor(nox_only$native_id)

size = 28 #per Blanco, the number of samples taken from each monitor
monitors = length(unique(nox_only$native_id)) #number of monitors in the data

# Initialize a dataframe to store the sample that is the same columns as the original nox_only df and enough rows for the desired sampling
sample_df <- data.frame(matrix(NA, nrow = size*monitors, ncol = ncol(nox_only)))
colnames(sample_df) <- colnames(nox_only)

set.seed(72) #seed for reproducability
# Initialize a row index for sample_df
row_index <- 1

for(i in 1:monitors) {
  data.i <- subset(nox_only, native_id_fact == unique(nox_only$native_id_fact)[i])
  sample_indices <- sample(1:nrow(data.i), size, replace = FALSE) # Take 'size' random samples from each monitor
  
  for(j in sample_indices) {
    sample_df[row_index, ] <- data.i[j, ]
    row_index <- row_index + 1
  }
}
sample_df$date <- as_datetime(sample_df$date) #make sure the date column stays in the right format in the new df. 

#head(sample_df) #Worked!

#Parameter name was a factor, so it only saved the number and not the name, but who cares (in the original df as imported before filtering 1 = NO2, 2 = NO, 3 = NOX)

#now writing it as function that will be easy to use

random_sample <- function(data, size, seed) {  #specify the data frame and the number of samples to take from each monitor
  data$native_id_fact <- factor(data$native_id)
  monitors <- length(unique(data$native_id)) # number of monitors in the data
  
  # Initialize a dataframe to store the sample that has the same columns as the original data and enough rows for the desired sampling
  sample_df <- data.frame(matrix(NA, nrow = size * monitors, ncol = ncol(data)))
  colnames(sample_df) <- colnames(data)
  
  set.seed(seed) # seed for reproducibility
  # Initialize a row index for sample_df
  row_index <- 1
  
  for(i in 1:monitors) {
    data.i <- subset(data, native_id == unique(data$native_id)[i])
    sample_indices <- sample(1:nrow(data.i), size, replace = FALSE) # Take 'size' random samples from each monitor
    
    for(j in sample_indices) {
      sample_df[row_index, ] <- data.i[j, ]
      row_index <- row_index + 1
    }
  }
  
  sample_df$date <- as_datetime(sample_df$date) # make sure the date column stays in the right format in the new df.
  
  return(sample_df)
}

test <- random_sample(nox_only, 28, 72)

#comparing test and sample_df to make sure they are the same

#check = sample_df$sample_measurement - test$sample_measurement
#max(check) #0
#min(check) #0
#The function works. 

```


```{r annual.avg, echo = FALSE, include = F}
#------------#Writing a function to aggregate data to the annual level--------------#

#function to take the annual average concentration of any dataframe
annual_avg <- function(data) {
  annual <- data %>% group_by(native_id) %>% summarise(annual_avg = mean(sample_measurement), log_avg = log(mean(sample_measurement)))
  return(annual)
}

#testing the function does what i want to to
test1 <- test %>% group_by(native_id) %>% summarise(annual_avg = mean(sample_measurement), log_avg = log(mean(sample_measurement)))

test2 <- annual_avg(test)

check = test1$annual_avg - test2$annual_avg
max(check) #0
min(check) #0
#The function works. 

check2 = test1$log_avg - test2$log_avg
max(check2) #0
min(check2) #0


```



```{r st.balanced, echo = F, include = F}

#taking a bunch of random samples for the short-term balanced data
# short term random samples meant to represent a balanced mobile monitoring campaign.
# Initialize the short-term balanced data frame
st_balanced <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(nox_only, 28, seed)
  annual_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  annual_avg_data$seed <- seed
  annual_avg_data$rep_number <- i
  
  # Append to dataframe
  st_balanced <- rbind(st_balanced, annual_avg_data)
}

```



```{r, echo = F, include = T}
#Creating Gold Standard Data - 
#summary stats for long-term gold and st balanced. 

#Gold Standard Data
long_term_gold <- annual_avg(nox_only)


```



 
```{r, summary stats, echo = F, include = F}

#writing a function for calculating summary statistics of an annual average df.
#function to make summary stats easier
sum_stats <- function(data) {
  summary <- data %>% summarise(min = min(annual_avg), 
                                            Q25 = quantile(annual_avg, probs = 0.25), 
                                            Q50 = quantile(annual_avg, probs = 0.5),
                                            Q75 = quantile(annual_avg, probs = 0.75),
                                            max = max(annual_avg),
                                            mean = mean(annual_avg),
                                            SD = sd(annual_avg)
                                            )
  return(summary)
  
}

#testing
#sum_stats(st_balanced)

#yay!!

```


```{r data.setup.for.map, echo = F, include = F}
#THis chuck sets up the data for the maps. It is independent of all our other df except for "nox_only" and "ca_covariates" so it can get moved anywhere after those variables are created. 
#get just the geographic variabiles needed from the covariate file
 
geom <- ca_covariates %>% dplyr::select("native_id", "lambert_x", "lambert_y", "County") #using lambert b/c I had trouble with the lat long for some reason

#get an annual dataset. (This is the same as the long_term gold) just repeated here, so the maps don't interfere with other analysis
annual_sf <- nox_only %>% 
  group_by(native_id) %>%
  summarise(annual_avg = mean(sample_measurement)) 

#getting a combined seasonal file that will allow for a faceted season map

seasonal_sf <- nox_only %>% group_by(native_id, season) %>%
  summarise(seasonal_avg = mean(sample_measurement))

#joing to the geometry
annual_sf <- left_join(annual_sf, geom, by = "native_id")
seasonal_sf <- left_join(seasonal_sf, geom, by = "native_id")


#convert the df to sf files (creating seperate sf for lambert and lat long prog to experiment maps)
annual_lam_sf <- st_as_sf(annual_sf, coords=c('lambert_x','lambert_y'), crs=lambert_proj)
annual_latlong_sf <- st_transform(annual_lam_sf, crs = latlong_proj)
seasonal_lam_sf <- st_as_sf(seasonal_sf, coords=c('lambert_x','lambert_y'), crs=lambert_proj)
seasonal_latlong_sf <- st_transform(seasonal_lam_sf, crs = latlong_proj)

```

```{r monitor.location.maps, echo = F, include = T}


#plotting annual averages as points on the monitors with color scale and testing lambert vs latlong crs

#adding lat_long limits and fixing up the map

#getting the boundaries of the extent 

lat_bbox <- st_bbox(annual_latlong_sf)
#print(lat_bbox)

#setting limits based on the bbox 
lon_limits <- c(-124.5, -114)
lat_limits <- c(32.5, 42)

#MAP 1

#same map but with the axis limits adjusted to frame CA
latlong <- ggplot(data = annual_latlong_sf , aes(fill = annual_avg)) + 
  ggspatial::annotation_map_tile("osm", alpha = 0.7) +  # Add the basemap first
  ggspatial::annotation_scale(location = "bl", width_hint = 0.3, unit_category = "imperial") +  # Scale in miles
  ggspatial::annotation_north_arrow(location = "tr", which_north = "true", height = unit(1, "cm"),
  width = unit(1, "cm"),) +  # North arrow
  geom_sf(size = 3, alpha = 0.6, shape = 21, color = "black") +  # Plot the monitors, adding alpha
  scale_fill_viridis_c(option = "plasma") +  # Apply the viridis color scale
  labs(title = "Fig. 1: CA Monitor Locations \nwith Annual Average NOx Concentration",
       fill = "NOx ppb",
       caption = "WGS84 projection") +
  coord_sf(xlim = lon_limits, ylim = lat_limits) + # Set the axis limits
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA)
  )

latlong

```

Figure 1: Shows the monitor locations across California. Colors indicate the annual average NOx concentration in ppb. As seen, the 69 monitor locations are widely distributed across a vast area, and there is not a large amount of variability in annual average concentration across monitors, with the majority of annual average concentration on the lower end of the scale. The greatest density of monitors as well as the greatest variation in annual averages occures in the Los Angeles metropolitan area, which motivated our sensitivity analysis in this region.


```{r socal.map, echo = F, include = T}
#zoomed in on S. CA to try and show more detail in the seasonal variation

#filter to so_cal only (can redefine what me mean by SCal later, but this is a maximum swath of counties)
counties <- c("Los Angeles", "Orange", "Riverside", "San Diego", "Ventura", "Santa Barbara", "San Bernardino") #list counties
socal_latlong_sf <- seasonal_latlong_sf %>% filter(County %in% counties)

#set new limits
socal_bbox = st_bbox(socal_latlong_sf)
#socal_bbox
socal_lon_limits <- c(-120.5, -116)
socal_lat_limits <- c(32.5, 35)

socal_seasons <- ggplot(data = socal_latlong_sf , aes(fill = seasonal_avg)) + 
  ggspatial::annotation_map_tile("osm", alpha = 0.7) +  # Add the basemap first
  geom_sf(size = 2.5, alpha = 0.7, shape = 21, color = "black") +  # Plot the monitors, adding alpha
  scale_fill_viridis_c(option = "plasma") +  # Apply the viridis color scale
  facet_wrap(~season) + 
  labs(title = "Fig. 2: Southern CA Monitor Locations with Seasonal Average NOx Concentration",
       fill = "NOx ppb",
       caption = "WGS84 projection") +
  coord_sf(xlim = socal_lon_limits, ylim = socal_lat_limits) + # Set the axis limits
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA)
  )

socal_seasons


```
Figure 2: Zooms in on Monitors in Southern CA and displays seasonal averages for each monitor. This figure includes more than just the 12 sensors we will use in our sensitivity analysis of LA county, but it does allow us to focus on the variability that might impact this restricted data set andthese seasonal models. In this figure, we can see that there is more variability in fall and winter when concentrations are overall higher than there is in spring and summer when concentrations are overall lower. 

```{r, include = F, echo = F}
#-----generate cross validation group for all analyses-----

# set the seed to make reproducible
set.seed(123)

# create vector of CV groups
  # first generate a vector with values 1:10 equal to the number of dataset rows
  # then randomize (with argument replace = FALSE)
CV_grp <- rep(1:10, length.out = nrow(long_term_gold)) %>% 
  sample(replace = FALSE)


# now append it to the cal_nox_clean data frame
long_term_gold <- mutate(long_term_gold, CV_grp = CV_grp)

```


```{r define get_MSE, include = F, echo = F}
#-----define get_MSE function-----

# This is a function to get the MSE, RMSE, MSE-based R2
get_MSE <- function(obs,pred) {
    # obs is the outcome variable
    # pred is the prediction from a model
     
    # mean of obs
    obs_avg <- mean(obs)
    
    # MSE of obs (for R2 denominator)
    MSE_obs <- mean((obs-obs_avg)^2)
    
    # MSE of predictions
    MSE_pred <- mean((obs - pred)^2)
    
    # compile output
    result <- c(RMSE = sqrt(MSE_pred),
                MSE_based_R2 = max(1 - MSE_pred / MSE_obs, 0),
                reg_based_R2 = max(cor(obs, pred)^2) #abbie added 12/9
                )
    
    # explicit return (optional)
    return(result)
}


```


```{r define.CV.function, include = F, echo = F}
#-----define CV function-----

do_CV <- function(data, id = "id", group = "group", formula) {
 
  lapply(unique(data[[group]]), function(this_group){
    
    # fit the "common" model to the training set (without this group)
    CV_lm <- lm(formula, data = data[data[[group]] != this_group,])
    
    # generate predictions for this group using training model
    data[data[[group]] == this_group,] %>%
      mutate(cvpreds = predict(CV_lm, newdata = .) %>% unname())
    
    # recombine data from all clusters and sort by ID column
    # note use of ".data[[ ]]" to return the value of variable id
  }) %>% bind_rows() %>% arrange(.data[[id]])
  
  # return the dataset (the last-evaluated object is always returned by default)
}

```


```{r covariates, include = F, echo = F}

mercer_cov <- ca_covariates %>%
  dplyr::select("native_id", "m_to_a1", "m_to_coast", "m_to_comm","pop_s05000", "ll_a1_s00050", "ll_a2_s00400", "ll_a3_s00400", "lu_industcomm_p05000", "County", "geometry")



```




```{r forward selection - gold standard, echo = F, include = F}

# forward, stepwise selection with common model covariates

#-----forward selection using interaction-----#
long_term_gold <- left_join(long_term_gold,mercer_cov, by = "native_id" )


null <- lm(log_avg ~ 1, data = long_term_gold)


covars_all <- str_subset(names(long_term_gold),"pop_|int_|open_|ll_|m_to_")

# B: now turn this into a formula for the full model in stepwise regression:
full <- as.formula(paste("log_avg ~ ", paste(covars_all, collapse= "+")))

# Using k=2 is comparable to standard AIC.
forwardreg_day <- step(null, 
                        scope = list(lower = null, upper = full), 
                        trace = 0, 
                        direction = "forward", 
                        k = 2)

covars_forward2 <- names(forwardreg_day$coefficients) %>%
  setdiff('(Intercept)')

covars_forward2 #different list than above when just run for summer



```


```{r cv in forward selection, warning = FALSE, message = FALSE, echo = F, include = F}
#gold-standard dataset for model selection

#-----model order and CV-----

# apply along length of the vector of names from forward selection
res2 <- lapply(seq_along(covars_forward2), function(i){
    
    # define the formula, updated to add a term each time
    fmla <- as.formula(
        paste("log_avg ~ + ", paste(covars_forward2[seq_len(i)], collapse = "+"))) 
    
    # in-sample model and estimates
    in_model <- lm(fmla, data = long_term_gold) 
    
    # out-of sample model and estimates
    out_ests <- do_CV(data = long_term_gold, id = "native_id", group = "CV_grp", fmla)  
    out_results <- get_MSE(out_ests$log_avg, out_ests$cvpreds)
    
    # compile results
   tibble(n_pred = i,
           covar = covars_forward2[i],
           in_RMSE = sqrt(mean(in_model$residuals^2)),
           in_R2 = summary(in_model)$r.squared,
           out_RMSE = out_results[["RMSE"]],
           out_R2 = out_results[["MSE_based_R2"]] 
           )
    
    }) %>% 
    bind_rows()

head(res2)

#I don't understand the purpose of this chunk? - KW



```


```{r bias.plots, echo = F, include = T}
#-----bias-var combined plots-----


#there are 4 terms in the model. 
#max(res2$out_RMSE[1:4])
#min(res2$out_RMSE[1:4])

y_lim <- 0.5

# create temporary dataframe for plot
temp2 <- res2 %>% 
  
  # make long dataframe
  pivot_longer(cols = c(ends_with("_RMSE"), ends_with("_R2")), 
               names_to = "Source_Estimate", 
               values_to = "value" ) %>%
  
  # separate the "Source" column for in and out of sample
  separate(col = Source_Estimate, into = c("Source", "Estimate") ) %>% 

  # set high RMSE values to NA, then filter out these values before plotting
  mutate(value = ifelse(Estimate == "RMSE" & value > y_lim, NA, value)) %>%
  filter(!is.na(value)) 


combined_plot2 <- ggplot(data = temp2) +
  geom_point(aes(x = n_pred, y = value, color = Source)) +
  geom_line(aes(x = n_pred, y = value, color = Source)) +
  xlab("Model Complexity (# of terms)") +
  ylab("") +
  labs(title = "Figure 3.0: Bias-Variance Trade-Off For 10-fold Cross-Validated CV Groups",
       subtitle = "Mercer Common Model Covariates") +
  #scale_x_continuous(breaks = c(seq(0, 63, 5))) +
  facet_wrap(~ Estimate, scales = "free_y", ncol = 1, strip.position = "right") +
  scale_color_discrete(labels = c('In-sample', 'Out-of-sample')) +
  theme_bw() 

#show plot
combined_plot2
```


**Adding Covariates from Mercer et al 2011 Common Model**
  The Mercer et al 2011 paper has the following covariates determined in its "common model" with which they assessed seasonal prediction models: D2A1 (distance to A1 road), A1_50m (length of A1 road in 50 m buffer), A23_400m (length of A2 or A3 roads within 400 m buffer), Pop_5k (total population within 5,000 m buffer), D2C (Distance to the coast in meters), Int_3k (intense land use within 3,000 m), D2Comm (distance to commercial land).

To assess the best fit of these covariates to our data, we ran forward, stepwise selection using k =2 as the penalty on the california nox covariates that matched these mercer et al (2011) covariates the best (a total of 8 covariates). This selection returned three covariates for use in our model. The covariates in our common model are:
  
"pop_s05000" - population within 5,000 m buffer, "m_to_coast" - Distance to the coast in meters, and "m_to_a1" - Distance to a A1 road in meters.  

Figure 3.0 shows that our out-of-sample performance peaks with the first three covariates included in forward selection and illustrates the appropriate selection of three, rather than four, covariates.. 

 
```{r, echo = F, include = F}
# build regression formula with common model covariates

covars_common <- c("pop_s05000" ,"m_to_coast" ,"m_to_a1")
frml <- as.formula(paste("log_avg ~", paste(covars_common, collapse = "+")) )

```


```{r, include = FALSE, echo = F}

#Running the prediction model on the lt_gold  - mercer

# in-sample predictions and fit summary
summary(lm_gold_standard <- lm(frml, data = long_term_gold))



```




```{r, echo = F, include = T}

#Doing CV on the gold_standard model
gold_CV <- do_CV(long_term_gold, id = "native_id", group = "CV_grp", formula = frml)

gold_MSE <- get_MSE(gold_CV$log_avg, gold_CV$cvpreds)

#gold_MSE

#       RMSE    MSE_based_R2 reg_based_R2 
 #  0.4644855    0.3346974    0.3397772 

#plotting the results

# get range for plot
r <- gold_CV %>% dplyr::select(log_avg, cvpreds) %>% range()


ggplot(data = gold_CV, aes(log_avg, cvpreds)) +
    geom_point(shape = "o", alpha = 0.8) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, aes(color = "red")) +
    geom_abline(aes(slope = 1, intercept = 0, color = "blue")) +
    labs(title = "Figure 4.0: Model predictions vs. observed ln(NOx)\nn Cross-validated", 
         subtitle = "Gold Standard Data",
         x = "Observed ln(NOx) (ln(ppb))",
         y = "Predicted ln(NOx) (ln(ppb))",
         caption = "RMSE = 0.464 MSE-based R2 = 0.335") + 
    scale_colour_manual(name='',
                      labels = c("1:1 Line", "Best-Fit Line"), 
                      values=c("blue", "red")) +
    theme_bw()


#model we have selected doesn't give the greatest predictions, but that might be fine. We changing the covariates in the model above won't necessitate changing any of the other code. 
```
Figure 4.0 shows the results from our cross validation of our gold standard data set and our common model. Using annual averages to predict across all sensors. The model's MSE-based $R^2$ is 0.335, using data from across the state of California. Our model does not perform exceedingly well. This is likely due to large distances over which the monitors are scatted and as well as limited variability in annual average concentrations. However, the purpose of the analysis is not to select and ideal model for predicting NOx across the state of California, but to assess the performance of short-term campagins in predicting long-term annual averages. We will compare our statewide balanced and seasonal campaigns against the observations from this model and compare their performance statistics relative to these performance statistics. In a true monitoring campaign for epidemiology studies, more time would be spent on model selection, but in this analysis we are more interested in commparitive between campaigns that overall model performance. 
     


```{r st_balanced, echo = F, include = F}

#Repeating above to compare predictions from the st_balanced to predictions from the gold standard

#need to group the st_balanced data by "campaign" before doing regression
st_balanced <- left_join(st_balanced, mercer_cov, by = "native_id" )

# and st balanced model
summary(lm_st_balanced <- lm(frml, data = st_balanced))

#apply the save CV groups as used in the gold standard model
st_balanced <- st_balanced %>% group_by(rep_number) %>% mutate(CV_grp = CV_grp) %>% ungroup()

#in sample-predictions
fitted_models = st_balanced %>% group_by(rep_number) %>% do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_balanced_CV <- st_balanced %>% group_by(rep_number) %>% do(model = do_CV(data = ., id = "native_id", group = "CV_grp", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_balanced_CV <- st_balanced_CV %>% rename(model_run = rep_number) #change column name so there isn't a conflict when I unnest it

st_balanced_CV <- st_balanced_CV %>%
  unnest(cols = c(model))

#appending the predictions from the gold_standard to the st_balanced_CV df for comparison
gold_preds <- gold_CV %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds = cvpreds)
st_balanced_CV <- left_join(st_balanced_CV, gold_preds, by = "native_id" )


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the gold-standard to do method a

gold_obs <- gold_CV %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs = log_avg)
st_balanced_CV <- left_join(st_balanced_CV, gold_obs, by = "native_id" )

# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_balanced_CV$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_balanced_CV %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_balanced_MSE <- data.frame(
  model_run = unique(st_balanced_CV$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_balanced_MSE_long <- st_balanced_MSE %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_balanced_CV %>% dplyr::select(gold_preds, cvpreds) %>% range()


st_balanced_scatter <- ggplot(data = st_balanced_CV, aes(x = gold_preds, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "Balanced Campaign \nvs Gold Predictions",
         caption = "1:1 line is dashed") +
    theme_bw()

st_balanced_scatter


#This needs some aesthetic adjustments: choose between dots or lines, pick a color scale that's helpful, get rid of the funky legend, but it's cool!!
    #looks a lot like the balanced predictions from the Blanco paper. 

#replicate blanco figure 3
st_balanced_boxplot <- ggplot(st_balanced_MSE_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Balanced Comparison") +
  theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0)
  )

st_balanced_boxplot

#This is also not bad. It will need some fancification, as well as adjustment when we start to add other campaigns, but it's an ok start. 

#Need to also add the in-sample R2, which I can do later.

#probably want to turn some of the above stuff into functions, because it is going to be a lot of cut and paste to do this for different data sets multiple times. 

# get range for plot
r <- st_balanced_CV %>% dplyr::select(log_avg, cvpreds) %>% range()


st_balanced_obs_v_preds <- ggplot(data = st_balanced_CV, aes(log_avg, cvpreds)) +
    geom_point(shape = "o", alpha = 0.8) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, aes(color = "red")) +
    geom_abline(aes(slope = 1, intercept = 0, color = "blue")) +
    labs(title = "Figure 4.0: Model predictions vs. observed ln(NOx)\n Cross-validated 10- fold", 
         subtitle = "Short-Term Balanced Data",
         x = "Observed ln(NOx) (ln(ppb))",
         y = "Predicted ln(NOx) (ln(ppb))") + 
    scale_colour_manual(name='',
                      labels = c("1:1 Line", "Best-Fit Line"), 
                      values=c("blue", "red")) +
    theme_bw() 
  

st_balanced_obs_v_preds

```



``` {r seasonal strata}
#----------create season-specific strata---------------

summer <- nox_only[nox_only$season == "Summer",]

fall <- nox_only[nox_only$season == "Fall",]

winter <- nox_only[nox_only$season == "Winter",] 

spring <- nox_only[nox_only$season == "Spring",]



```



```{r seasonal.avg, echo = FALSE, include = F, eval = F}
#Writing a function to aggregate data to the seasonal level -seasonal dataset

#function to take the annual average concentration of any dataframe
seasonal_avg <- function(data) {
  seasonal <- data %>% group_by(native_id) %>% summarise(seasonal_avg = mean(sample_measurement), log_avg = log(mean(sample_measurement)))
  return(seasonal)
}

#I set evalu = F on this. I am going to use the annual_avg() function, because I want the column names to all be names the same thing when I bind them together for the plots. 

```


```{r seasonal datasets2, echo = F, include = F}


#random_sample() (data, size, seed)
#----------summer----------------------------------------------
# Initialize the short-term balanced data frame
st_summer <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(summer, 28, seed)
  summer_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  summer_avg_data$seed <- seed
  summer_avg_data$rep_number <- i
  
  # Append to st summer df
  st_summer <- rbind(st_summer, summer_avg_data)
}
summer_state_var<-ggplot(st_summer, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Figure 5.0: Summer Short Term Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor")


#----------fall----------------------------------------------
# Initialize the short-term balanced data frame
st_fall <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(fall, 28, seed)
  fall_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  fall_avg_data$seed <- seed
  fall_avg_data$rep_number <- i
  
  # Append to st fall df
  st_fall <- rbind(st_fall, fall_avg_data)
}
fall_state_var<-ggplot(st_fall, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Fall Short Term Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor")

#------------Winter--------------------------------------------
st_winter <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(winter, 28, seed)
  winter_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  winter_avg_data$seed <- seed
  winter_avg_data$rep_number <- i
  
  # Append to st winter df
  st_winter <- rbind(st_winter, winter_avg_data)
}
winter_state_var<-ggplot(st_winter, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Winter Short Term Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor")


#------------Spring--------------------------------------------
st_spring <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(spring, 28, seed)
  spring_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  spring_avg_data$seed <- seed
  spring_avg_data$rep_number <- i
  
  # Append to st spring df
  st_spring <- rbind(st_spring, spring_avg_data)
}

spring_state_var<- ggplot(st_spring, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Spring Short Term Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor")


# this should be the same length as the st_balanced dataframe
#True - AG

#Looks good! Only suggestion is that we just use the annual_avg() function and not add a season_avg() function, because other than the name, they produce the same results. KW

#if you think these plots are helpful, then we should spend some time fixing them up. I think they are mostly just to help us visualize the data, and can be left out (or put in appendix without the grid arrange)KW

#moved to the appendix - AG

```

```{r, results='hide',include=FALSE,echo=FALSE}
#summary statistics
lt_gold_sum <- long_term_gold %>% summarise(min = min(annual_avg), 
                                            Q25 = quantile(annual_avg, probs = 0.25), 
                                            Q50 = quantile(annual_avg, probs = 0.5),
                                            Q75 = quantile(annual_avg, probs = 0.75),
                                            max = max(annual_avg),
                                            mean = mean(annual_avg),
                                            SD = sd(annual_avg)
                                            )
#lt_gold_sum
      #This matches the number reported in Blanco supplement S5

#summary stats for st_balanced
st_balanced_sum <- st_balanced %>% summarise(min = min(annual_avg), 
                                            Q25 = quantile(annual_avg, probs = 0.25), 
                                            Q50 = quantile(annual_avg, probs = 0.5),
                                            Q75 = quantile(annual_avg, probs = 0.75),
                                            max = max(annual_avg),
                                            mean = mean(annual_avg),
                                            SD = sd(annual_avg)
                                            )
#st_balanced_sum #this would not be expected to match Blanco because we used different seeds
#summarizing the data from the random samples
names <- c("Gold Standard (LT)", "ST Balanced")

kable(cbind(names , rbind(lt_gold_sum, st_balanced_sum)), digits = 2,
      caption = "Table 1.0: Balanced, Long-term & Short-term Sampling Campaign Summaries") %>% kable_styling()

```


```{r, echoo = F, include = T}
#summary stats for each season

st_summer_sum <- sum_stats(st_summer) 

#st_summer_sum

st_fall_sum <- sum_stats(st_fall) 
#st_fall_sum 

st_winter_sum <- sum_stats(st_winter)
#st_winter_sum 

st_spring_sum <- sum_stats(st_spring) 
#st_spring_sum



Datasets <- c("summer (st)", "fall (st)", "winter (st)", "spring (st)", "balanced (st)", "gold standard (lt)")
kable(cbind(Datasets , rbind(st_summer_sum, st_fall_sum, st_winter_sum, st_spring_sum, st_balanced_sum, lt_gold_sum)), digits = 2,
      caption = " Table 1.0: Sampling Campaigns NOx (ppb) Summaries") %>%
  kable_styling()

```
Table 2.0 summarizes the spread and range of the data randomly sampled for each data set in our statewide analysis and the gold standard dataset. The Fall and Winter data sets have nearly double the variation as Spring and Summer data sets. As we would expect, the balanced data set has variability in the middle between these four seasonal data sets. As we would expect, the balanced and gold standard data sets have a standard deviation in the middle of our seasonal data sets. 


```{r season.convariates2, echo = F, include = F}
#--------------------------Common Model on Seasonal Data sets------------------#
#adding the covariates to the gold standard data to do some predictions
st_summer <-  left_join(st_summer,mercer_cov, by = "native_id" )
st_fall <-  left_join(st_fall,mercer_cov, by = "native_id" )
st_winter <-  left_join(st_winter,mercer_cov, by = "native_id" )
st_spring <-  left_join(st_spring,mercer_cov, by = "native_id" ) 


```


```{r, include = FALSE, echo = F}

#running the regression for each model 

#we don't need to include these summaries in the html
# in-sample predictions and fit summary
summary(lm_st_summer <- lm(frml, data = st_summer))
summary(lm_st_fall <- lm(frml, data = st_fall))
summary(lm_st_winter <- lm(frml, data = st_winter))
summary(lm_st_spring <- lm(frml, data = st_spring)) 


```


#----------comparing seasonal short term campaigns with Gold Standard data----------------

```{r summer vs. gold, echo = F, include = F}
#------------seasonal comparison with gold standard-------------

#-----------------------------------SUMMER--------------------------------------
#need to group the st_summer data by "campaign" before doing regression

#apply the save CV groups as used in the gold standard model
st_summer <- st_summer %>% group_by(rep_number) %>% mutate(CV_grp = CV_grp) %>% ungroup()

#in sample-predictions
fitted_models = st_summer %>% group_by(rep_number) %>% do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_summer_CV <- st_summer %>% group_by(rep_number) %>% do(model = do_CV(data = ., id = "native_id", group = "CV_grp", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_summer_CV <- st_summer_CV %>% rename(model_run = rep_number) #change column name so there isn't a conflict when I unnest it

st_summer_CV <- st_summer_CV %>%
  unnest(cols = c(model))

#---compare st_summer with gold-----------------------------------------#
#appending the predictions from the balanced short term campaign to the st_summer_CV df for comparison
gold_preds <- gold_CV %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds = cvpreds)  #select "model run" as well as native_id
st_summer_CV <- left_join(st_summer_CV, gold_preds, by = "native_id" ) #to have unique identifiers we need both native id and model run. AG


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the balanced short term campaign to do method a

gold_obs <- gold_CV %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs = log_avg)
st_summer_CV <- left_join(st_summer_CV, gold_obs, by = "native_id" ) #to have unique identifiers we need both native id and model run. AG

# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_summer_CV$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_summer_CV %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_summer_MSE <- data.frame(
  model_run = unique(st_summer_CV$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_summer_MSE_long <- st_summer_MSE %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_summer_CV %>% dplyr::select(gold_preds, cvpreds) %>% range()


st_summer_scatter <- ggplot(data = st_summer_CV, aes(x = gold_preds, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "Summer Campaign \nvs Gold Predictions", 
         caption = "1:1 line is dashed") +
    theme_bw() +
  theme(plot.caption = element_text(hjust = 0)
  )
  
st_summer_scatter


#This needs some aesthetic adjustments: choose between dots or lines, pick a color scale that's helpful, get rid of the funky legend, but it's cool!!
    #looks a lot like the balanced predictions from the Blanco paper. 

#replicate blanco figure 3
st_summer_boxplot <-  ggplot(st_summer_MSE_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Summer Comparison") +
  theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0)
  )


  st_summer_boxplot


```



```{r fall vs gold, echo = F, include = F}
#------------seasonal comparison with gold standard-------------

#-----------------------------------fall--------------------------------------
#need to group the st_fall data by "campaign" before doing regression

#apply the save CV groups as used in the gold standard model
st_fall <- st_fall %>% group_by(rep_number) %>% mutate(CV_grp = CV_grp) %>% ungroup()

#in sample-predictions
fitted_models = st_fall %>% group_by(rep_number) %>% do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_fall_CV <- st_fall %>% group_by(rep_number) %>% do(model = do_CV(data = ., id = "native_id", group = "CV_grp", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_fall_CV <- st_fall_CV %>% rename(model_run = rep_number) #change column name so there isn't a conflict when I unnest it

st_fall_CV <- st_fall_CV %>%
  unnest(cols = c(model))

#---compare st_fall with gold-----------------------------------------#
#appending the predictions from the balanced short term campaign to the st_fall_CV df for comparison
gold_preds <- gold_CV %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds = cvpreds)  #select "model run" as well as native_id
st_fall_CV <- left_join(st_fall_CV, gold_preds, by = "native_id" ) #to have unique identifiers we need both native id and model run. AG


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the balanced short term campaign to do method a

gold_obs <- gold_CV %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs = log_avg)
st_fall_CV <- left_join(st_fall_CV, gold_obs, by = "native_id" ) #to have unique identifiers we need both native id and model run. AG

# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_fall_CV$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_fall_CV %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_fall_MSE <- data.frame(
  model_run = unique(st_fall_CV$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_fall_MSE_long <- st_fall_MSE %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_fall_CV %>% dplyr::select(gold_preds, cvpreds) %>% range()


st_fall_scatter <-ggplot(data = st_fall_CV, aes(x = gold_preds, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "fallCampaign \nvs Gold Predictions", 
         caption = "1:1 line is dashed") +
    theme_bw() +
  theme(plot.caption = element_text(hjust = 0)
  )

st_fall_scatter


#This needs some aesthetic adjustments: choose between dots or lines, pick a color scale that's helpful, get rid of the funky legend, but it's cool!!
    #looks a lot like the balanced predictions from the Blanco paper. 

#replicate blanco figure 3
st_fall_boxplot <-  ggplot(st_fall_MSE_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Fall Comparison") +
  theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0)
  )

st_fall_boxplot
  

```



```{r winter v gold, echo = F, include = F}
#------------seasonal comparison with  and gold standard-------------

#-----------------------------------winter--------------------------------------
#need to group the st_winter data by "campaign" before doing regression

#apply the save CV groups as used in the gold standard model
st_winter <- st_winter %>% group_by(rep_number) %>% mutate(CV_grp = CV_grp) %>% ungroup()

#in sample-predictions
fitted_models = st_winter %>% group_by(rep_number) %>% do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_winter_CV <- st_winter %>% group_by(rep_number) %>% do(model = do_CV(data = ., id = "native_id", group = "CV_grp", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_winter_CV <- st_winter_CV %>% rename(model_run = rep_number) #change column name so there isn't a conflict when I unnest it

st_winter_CV <- st_winter_CV %>%
  unnest(cols = c(model))

#---compare st_winter with gold-----------------------------------------#
#appending the predictions from the balanced short term campaign to the st_winter_CV df for comparison
gold_preds <- gold_CV %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds = cvpreds) 
st_winter_CV <- left_join(st_winter_CV, gold_preds, by = "native_id" ) 


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the balanced short term campaign to do method a

gold_obs <- gold_CV %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs = log_avg)
st_winter_CV <- left_join(st_winter_CV, gold_obs, by = "native_id" ) 

# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_winter_CV$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_winter_CV %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_winter_MSE <- data.frame(
  model_run = unique(st_winter_CV$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_winter_MSE_long <- st_winter_MSE %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_winter_CV %>% dplyr::select(gold_preds, cvpreds) %>% range()


st_winter_scatter <-ggplot(data = st_winter_CV, aes(x = gold_preds, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "winter  Campaign \nvs Gold Predictions", 
         caption = "1:1 line is dashed") +
    theme_bw() +
  theme(plot.caption = element_text(hjust = 0)
  )

st_winter_scatter

#This needs some aesthetic adjustments: choose between dots or lines, pick a color scale that's helpful, get rid of the funky legend, but it's cool!!
    #looks a lot like the balanced predictions from the Blanco paper. 

#replicate blanco figure 3
st_winter_boxplot <- ggplot(st_winter_MSE_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Winter Comparison") +
  theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0)
  )

st_winter_boxplot

```



```{r spring v gold, echo = F, include = F}
#------------seasonal comparison with st balanced and gold standard-------------

#-----------------------------------spring--------------------------------------
#need to group the st_spring data by "campaign" before doing regression

#apply the save CV groups as used in the gold standard model
st_spring <- st_spring %>% group_by(rep_number) %>% mutate(CV_grp = CV_grp) %>% ungroup()

#in sample-predictions
fitted_models = st_spring %>% group_by(rep_number) %>% do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_spring_CV <- st_spring %>% group_by(rep_number) %>% do(model = do_CV(data = ., id = "native_id", group = "CV_grp", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_spring_CV <- st_spring_CV %>% rename(model_run = rep_number) #change column name so there isn't a conflict when I unnest it

st_spring_CV <- st_spring_CV %>%
  unnest(cols = c(model))

#---compare st_spring with gold-----------------------------------------#
#appending the predictions from the balanced short term campaign to the st_spring_CV df for comparison
gold_preds <- gold_CV %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds = cvpreds)  #select "model run" as well as native_id
st_spring_CV <- left_join(st_spring_CV, gold_preds, by = "native_id" ) #to have unique identifiers we need both native id and model run. AG


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the balanced short term campaign to do method a

gold_obs <- gold_CV %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs = log_avg)
st_spring_CV <- left_join(st_spring_CV, gold_obs, by = "native_id" )

# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_spring_CV$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_spring_CV %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_spring_MSE <- data.frame(
  model_run = unique(st_spring_CV$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_spring_MSE_long <- st_spring_MSE %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_spring_CV %>% dplyr::select(gold_preds, cvpreds) %>% range()


st_spring_scatter <-ggplot(data = st_spring_CV, aes(x = gold_preds, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "Spring Campaign \nvs Gold Predictions",
         caption = "1:1 line is dashed") +
    theme_bw() +
  theme(plot.caption = element_text(hjust = 0)
  )

st_spring_scatter


#This needs some aesthetic adjustments: choose between dots or lines, pick a color scale that's helpful, get rid of the funky legend, but it's cool!!
    #looks a lot like the balanced predictions from the Blanco paper. 

#replicate blanco figure 3
st_spring_boxplot <- ggplot(st_spring_MSE_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Spring Comparison") +
  theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0)
  )

st_spring_boxplot

```



In Figure X.X, the box plot plots, the Balanced campaign is the only short term campaign with a reasonable R squared and reasonable performance in comparison witht he gold standard. The Agreement between the gold standard and the short term seasonal campaigns is very low, with R squared values near zero (pink). The blue boxplots show the cross validated performance statistics for each campaign, which we see hovering around 0.3 with all campaigns.

In figure X.X., the scatter plots show trends in over- or under-prediction by each short term campaign when compared to the gold standard predictions. We can see that the spring and summer models under predict ln(NOx) while the winter and fall models over-predict ln(NOx). The Balanced short term campaign doesn't appear to have a visible trend in over- or under-predicting ln(NOX) concentrations and seems to agree well with the gold standard predictions, with plotted predictions falling right along the 1:1 line. 



```{r, echo = F, include = T, fig.height = 7, fig.width = 7, warning = F, message = F}

#trying to make a better plot

#add a season column to each one
st_spring_CV <- st_spring_CV %>% mutate(season = "Spring")
st_summer_CV <- st_summer_CV %>% mutate(season = "Summer")
st_fall_CV <- st_fall_CV %>% mutate(season = "Fall")
st_winter_CV <- st_winter_CV %>% mutate(season = "Winter")
st_balanced_CV <- st_balanced_CV %>% mutate(season = "Balanced")

#bind the df into one
st_all_CV <- rbind(st_balanced_CV, st_spring_CV, st_summer_CV, st_fall_CV, st_winter_CV)

#factor to control the order
st_all_CV <- st_all_CV %>%
  mutate(season = factor(season, levels = c("Spring", 
                                            "Summer", 
                                            "Fall", 
                                            "Winter", 
                                            "Balanced")))


all_scatter <-ggplot(data = st_all_CV, aes(x = gold_preds, y = cvpreds, color = as.factor(model_run))) +
    #geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) + #commented out the points to make the plot a little cleaner. 
    lims(x= r, y = r) +
    geom_smooth(method = 'lm', se = F, alpha = 0.4, show.legend = FALSE, linewidth = 0.3) +
    facet_wrap(~season, scales = "free") + 
    geom_abline(linetype = "dashed", slope = 1, intercept = 0, color = "darkgray") +
    labs(title = "Fig. X: Best fit lines of cross-validated short-term predictions for 30 campaigns\nvs. gold standard predictions for ln(NOx)", 
         subtitle = "Statewide",
         x = "Gold standard predicted ln(NOx) (ln(ppb))",
         y = "Campaign Predicted ln(NOx) (ln(ppb))",
         caption = "1:1 line is dashed") 
    theme_bw()

all_scatter

```
This figure shows the performance of each of the 30 campaigns in within each short-term sampling design and compare the campaign predictions to the predictions from the gold standard. The balanced design performs well. Summer and spring consistently over predict (predict higher values than the gold standard), and fall and winter underpredict compared to the gold standard. *check my understanding on under andd over prediction*

The scatter plots in Figure 7.0 and the box plots in figure 6.0 compare the performance of the common model on the seasonal and balanced short term data sets, as well as a comparison of the predicted NOx concentrations compared to the gold standard predictions. 

All of the seasonal models perform poorly compared to the gold standard model, but with different trends in their predictions. The Spring and Summer models under-predicted *I think it's over predict. check my thinking, but if you look at the plots, the gold standard would have predicted 2, the summmer model predicts 2.5* NOx compared to the gold standard, while the winter and fall models over-predicted in comparison with the gold standard predictions. 

The balanced model and data set performed the most similar to the gold standard model and data set, with no clear over- or under- predictions of NOx and a comparable R squared and RMSE. 

#tabulate RMSE and R sqaured for all campaigns:

```{r table of prediction stats, echo = F, include = T}
#averages the rmse and r squared across all 30 repeats of the random sampling for all campaigns

campaigns <- c("Summer", "Fall", "Winter", "Spring", "Balanced")
RMSE <- c(mean(st_summer_MSE$RMSE_a), mean(st_fall_MSE$RMSE_a), mean(st_winter_MSE$RMSE_a), mean(st_spring_MSE$RMSE_a), mean(st_balanced_MSE$RMSE_a)) %>% round(digits = 3)
MSE_R2 <- c(mean(st_summer_MSE$MSE.R2_a), mean(st_fall_MSE$MSE.R2_a), mean(st_winter_MSE$MSE.R2_a), mean(st_spring_MSE$MSE.R2_a), mean(st_balanced_MSE$MSE.R2_a)) %>% round(digits = 3)
Reg_R2 <- c(mean(st_summer_MSE$reg.R2_a), mean(st_fall_MSE$reg.R2_a), mean(st_winter_MSE$reg.R2_a), mean(st_spring_MSE$reg.R2_a), mean(st_balanced_MSE$reg.R2_a)) %>% round(digits = 3)

kable(rbind(campaigns, RMSE, MSE_R2, Reg_R2),
      digits = 3,
      caption = "Table 3.0: Performance Statistics of Seasonal and Balanced Sampling Campaigns - Method A") %>% kable_styling()

```
Based on the R squared and RMSE values in Table 3.0 we, we see that the balanced short term data set performed the best out of the short term statewide campaigns. 

```{r, echo = F, include = T, fig.height = 7, fig.width = 7}

#bind the MSE df together for a combined boxplot
#add a season column to each one
st_spring_MSE_long <- st_spring_MSE_long %>% mutate(season = "Spring")
st_summer_MSE_long <- st_summer_MSE_long %>% mutate(season = "Summer")
st_fall_MSE_long <- st_fall_MSE_long %>% mutate(season = "Fall")
st_winter_MSE_long <- st_winter_MSE_long %>% mutate(season = "Winter")
st_balanced_MSE_long <- st_balanced_MSE_long %>% mutate(season = "Balanced")

st_all_MSE_long <- rbind(st_spring_MSE_long, st_summer_MSE_long, st_fall_MSE_long, st_winter_MSE_long, st_balanced_MSE_long)

#factor to control the order
st_all_MSE_long <- st_all_MSE_long %>%
  mutate(season = factor(season, levels = c("Spring", 
                                            "Summer", 
                                            "Fall", 
                                            "Winter", 
                                            "Balanced")))

#replicate blanco figure 3
st_all_boxplot <- ggplot(st_all_MSE_long, aes(x = season, y = value, color = season)) +
  geom_boxplot() +
  geom_hline(data = data.frame(method = c("a", "b"), value = gold_MSE[[1]], metric = "RMSE"), aes(yintercept = value, linetype = "dotted"), show.legend = T) +
  geom_hline(data = data.frame(method = c("a", "b"), value = gold_MSE[[2]], metric = "MSE.R2"), aes(yintercept = value, linetype = "dashed"), show.legend = T) +
  geom_hline(data = data.frame(method = c("a", "b"), value = gold_MSE[[3]], metric = "reg.R2"), aes(yintercept = value, linetype = "dotdash"), show.legend = T) +
  facet_grid(metric~method, scales = "free") +
  labs(title = "RMSE and R2 Comparison",
       caption = "Method A = Short-term campaign predictions compared to gold-standard obs\nMethod B = Short-term campaign predictions compared to same campaign observations\nHorizontal lines show the gold-standard performance", 
       x = "Short-term Campaign",
       y = "Value",
       color = "Season") +
  scale_linetype_manual(name='',
                      labels = c("Gold standard RMSE", "Gold Standard MSE-R2", "Gold Standard Reg. R2"), 
                      values=c("dotted", "dashed", "dotdash")) +
  theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()# Remove x-axis text labels
  )

st_all_boxplot



```





``` {r LA strat, echo = F, include = F}
#----------create LA-specific dataset---------------

la <-left_join(nox_only, mercer_cov, by = "native_id") %>% filter(County == "Los Angeles")

#head(la)

# fixed this so that it is more similar to the other df and takes a lot less time than using all the ca covarites. 


```


```{r long term gold standard la dataset, echo = F, include = F}

#Gold Standard Data for LA
long_term_gold_la <- annual_avg(la)


```





```{r short term LA dataset, echo = F, include = F}
#----------generate short-term LA dataset---------------

#random_sample() (data, size, seed)
# Initialize the short-term balanced data frame
st_la <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(la, 28, seed)
  la_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  la_avg_data$seed <- seed
  la_avg_data$rep_number <- i
  
  # Append to st summer df
  st_la <- rbind(st_la, la_avg_data)
}
balanced_LA<- ggplot(st_la, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Figure 8.0:Los Angeles County Short Term Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor")

```


```{r summarize LA short term dataset, include = F, echo = F}
#quick summary of long-term and short-term LA datasets

#summary stats for st_la
st_la_sum <- sum_stats(st_la) 

#summary stats for lt_gold_sum_la
lt_gold_sum_la <- sum_stats(long_term_gold_la) 


#combined table short and long term LA (balanced datasets)

names_la <- c("short-term balanced", "long-term balanced")
#kable(cbind(names_la , rbind(st_la_sum, lt_gold_sum_la)), digits = 2,
#      caption = " Short- and Long-Term Balanced Sampling Campaigns - NOx Summaries (Los Angeles) ") %>%
#  kable_styling()

# no need to have these all throughut the paper - AG

```

```{r la.convariates, include = F, echo = F}
#--------------------------Common Model on LA datasets------------------#

#adding the covariates to do some predictions
st_la<-  left_join(st_la,mercer_cov, by = "native_id")

long_term_gold_la <- left_join(long_term_gold_la,mercer_cov, by = "native_id")

```


```{r, , include = F, echo = F}
#running a regression model using common model covariates and LA data sets

# in-sample predictions and fit summary
summary(lm_st_la <- lm(frml, data = st_la)) #short term balanced data set

summary(lm_gold_la <- lm(frml, data = long_term_gold_la)) #short term balanced data set

```




```{r LOO cross validation groups, include = F, echo = F}
# cross-validation groups that can be used for all the data- leave-one-out to avoid too small of groups

CV_grp_la <- rep(1:12, length.out = nrow(long_term_gold_la)) %>% 
  sample(replace = FALSE)

# now append it to the cal_nox_clean data frame
long_term_gold_la <- mutate(long_term_gold_la, CV_grp_la = CV_grp_la)


#doing gold cv within LA dataset for long term gold standard model
gold_CV_la <- do_CV(long_term_gold_la, id = "native_id", group = "CV_grp_la", formula = frml)

gold_MSE_la <- get_MSE(gold_CV_la$log_avg, gold_CV_la$cvpreds)


# now make groups for short term data sets
CV_grp_st_la <- rep(1:360, length.out = nrow(st_la)) %>% 
  sample(replace = FALSE)




```




```{r la short-term balanced versus gold standard, include = F, echo = F}
#------------LAshort-term balanced versus gold standard-------------

#need to group the st_la data by "campaign" before doing regression

#apply the save CV groups as used in the gold standard model
st_la <- st_la %>% #group_by(rep_number) %>% #abbie had to remove this to have the leave one out groups append. 
  mutate(CV_grp_st_la = CV_grp_st_la) %>% 
  ungroup()

#in sample-predictions
fitted_models = st_la %>% group_by(rep_number) %>% 
  do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_la_CV <- st_la %>% group_by(rep_number) %>%
  do(model = do_CV(data = ., id = "native_id", group = "CV_grp_st_la", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_la_CV <- st_la_CV %>% rename(model_run = rep_number)

st_la_CV <- st_la_CV %>%
  unnest(cols = c(model))

#---compare st_la with gold la--------

#appending the predictions from the balanced short term campaign to the st_la_CV df for comparison
gold_preds_la <- gold_CV_la %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds_la = cvpreds)  
st_la_CV <- left_join(st_la_CV, gold_preds_la, by = "native_id" ) 


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the balanced short term campaign to do method a

gold_obs_la <- gold_CV_la %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs_la = log_avg)
st_la_CV <- left_join(st_la_CV, gold_obs_la, by = "native_id" )

# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_la_CV$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_la_CV %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs_la, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_la_MSE <- data.frame(
  model_run = unique(st_la_CV$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_la_MSE_long <- st_la_MSE %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_la_CV %>% dplyr::select(gold_preds_la, cvpreds) %>% range()


balanced_scatter_la<-ggplot(data = st_la_CV, aes(x = gold_preds_la, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "Short term Balanced \nvs Gold Standard Predictions", 
         caption = "1:1 line is dashed") +
    theme_bw()

#replicate blanco figure 3
balanced_boxplot_la<- ggplot(st_la_MSE_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Balanced Comparison") +
  theme_bw()

#NOTE: can someone please confirm that I got the captions right here? -ck Looks OK. We need to update the GOLD mse to just la though. Can do this is the final plot. 

```




``` {r seasonal strat}
#----------create season-specific strata---------------
summer_la <- la[la$season == "Summer",]

fall_la <- la[la$season == "Fall",]

winter_la <- la[la$season == "Winter",] 

spring_la <- la[la$season == "Spring",]



```


```{r seasonal datasets}


#random_sample() (data, size, seed)
#----------summer_la----------------------------------------------
# Initialize the short-term balanced data frame
st_summer_la <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(summer_la, 28, seed)
  summer_la_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  summer_la_avg_data$seed <- seed
  summer_la_avg_data$rep_number <- i
  
  # Append to st summer_la df
  st_summer_la <- rbind(st_summer_la, summer_la_avg_data)
}
summer_LA<- ggplot(st_summer_la, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Figure X: summer_la Short Term Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor")


#----------fall_la----------------------------------------------
# Initialize the short-term balanced data frame
st_fall_la <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(fall_la, 28, seed)
  fall_la_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  fall_la_avg_data$seed <- seed
  fall_la_avg_data$rep_number <- i
  
  # Append to st fall_la df
  st_fall_la <- rbind(st_fall_la, fall_la_avg_data)
}
fall_LA<-ggplot(st_fall_la, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Figure X: fall_la Short Term Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor")

#------------winter_la--------------------------------------------
st_winter_la <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(winter_la, 28, seed)
  winter_la_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  winter_la_avg_data$seed <- seed
  winter_la_avg_data$rep_number <- i
  
  # Append to st winter_la df
  st_winter_la <- rbind(st_winter_la, winter_la_avg_data)
}
winter_LA<-ggplot(st_winter_la, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Figure X: winter_la Short Term Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor")


#------------spring_la--------------------------------------------
st_spring_la <- data.frame()

n_reps <- 30  # number of random df simulations to run

for (i in 1:n_reps) {
  seed <- i * 4  # generating a different seed for each simulation
  sampled_data <- random_sample(spring_la, 28, seed)
  spring_la_avg_data <- annual_avg(sampled_data)
  
  # Add seed and rep_number columns
  spring_la_avg_data$seed <- seed
  spring_la_avg_data$rep_number <- i
  
  # Append to st spring_la df
  st_spring_la <- rbind(st_spring_la, spring_la_avg_data)
}

spring_LA<-ggplot(st_spring_la, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Figure X:spring_la Short Term Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor")






```

```{r, summary stat, echo = F, include = T}
#summary stats for LA

st_summer_la_sum <- sum_stats(st_summer_la) 

#st_summer_la_sum

st_fall_la_sum <- sum_stats(st_fall_la) 
#st_fall_la_sum 

st_winter_la_sum <- sum_stats(st_winter_la)
#st_winter_la_sum 

st_spring_la_sum <- sum_stats(st_spring_la) 
#st_spring_la_sum



Datasets <- c("Spring","Summer", "Fall", "Winter", "Balanced", "Gold-standard")
kable(cbind(Datasets , rbind(st_spring_la_sum, st_summer_la_sum, st_fall_la_sum, st_winter_la_sum, st_la_sum, lt_gold_sum_la)), digits = 2,
      caption = " Table X.0: Sampling Campaign NOx (ppb) Summaries for Los Angeles County") %>%
  kable_styling()


```


```{r season.convariates}
#--------------------------Common Model on Seasonal Data sets------------------#
#adding the covariates to the gold standard data to do some predictions
st_summer_la <-  left_join(st_summer_la,mercer_cov, by = "native_id" )
st_fall_la <-  left_join(st_fall_la,mercer_cov, by = "native_id" )
st_winter_la <-  left_join(st_winter_la,mercer_cov, by = "native_id" )
st_spring_la <-  left_join(st_spring_la,mercer_cov, by = "native_id" )


```



```{r, include = FALSE}

# in-sample predictions and fit summary
summary(lm_st_summer_la <- lm(frml, data = st_summer_la))
summary(lm_st_fall_la <- lm(frml, data = st_fall_la))
summary(lm_st_winter_la <- lm(frml, data = st_winter_la))
summary(lm_st_spring_la <- lm(frml, data = st_spring_la)) 


```





```{r summer vs gold}
#------------seasonal comparison with gold standard in LA county-------------

#-----------------------------------SUMMER--------------------------------------
#need to group the st_summer data by "campaign" before doing regression

#apply the save CV groups as used in the gold standard model
sst_summer_la <- st_summer_la %>%
  mutate(CV_grp_st_la = CV_grp_st_la) %>% ungroup()

#in sample-predictions
fitted_models = sst_summer_la %>% group_by(rep_number) %>% do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_summer_CV_la <- sst_summer_la %>% group_by(rep_number) %>% do(model = do_CV(data = ., id = "native_id", group = "CV_grp_st_la", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_summer_CV_la <- st_summer_CV_la %>% rename(model_run = rep_number) #change column name so there isn't a conflict when I unnest it

st_summer_CV_la <- st_summer_CV_la %>%
  unnest(cols = c(model))

#---compare st_summer with gold-----------------------------------------#
#appending the predictions from the balanced short term campaign to the st_summer_CV df for comparison
gold_preds_la <- gold_CV_la %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds_la = cvpreds)  #select "model run" as well as native_id
st_summer_CV_la <- left_join(st_summer_CV_la, gold_preds_la, by = "native_id" ) 


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the balanced short term campaign to do method a

gold_obs_la <- gold_CV_la %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs_la = log_avg)
st_summer_CV_la <- left_join(st_summer_CV_la, gold_obs_la, by = "native_id" ) 
# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_summer_CV_la$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_summer_CV_la %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs_la, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_summer_MSE_la <- data.frame(
  model_run = unique(st_summer_CV_la$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_summer_MSE_la_long <- st_summer_MSE_la %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_summer_CV_la %>% dplyr::select(gold_preds_la, cvpreds) %>% range()


summer_scatter_la<-ggplot(data = st_summer_CV_la, aes(x = gold_preds_la, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "Summer Campaign \nvsGold Standard Predictions", 
         caption = "1:1 line is dashed") +
    theme_bw()


#This needs some aesthetic adjustments: choose between dots or lines, pick a color scale that's helpful, get rid of the funky legend, but it's cool!!
    #looks a lot like the balanced predictions from the Blanco paper. 

#replicate blanco figure 3
summer_boxplot_la<- ggplot(st_summer_MSE_la_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Summer Comparison") +
  theme_bw()


  


```


```{r}
#------------seasonal comparison with gold standard in LA county-------------

#-----------------------------------FALL--------------------------------------
#need to group the st_summer data by "campaign" before doing regression

#apply the save CV groups as used in the gold standard model
sst_fall_la <- st_fall_la %>%
  mutate(CV_grp_st_la = CV_grp_st_la) %>% ungroup()

#in sample-predictions
fitted_models = sst_fall_la %>% group_by(rep_number) %>% do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_fall_CV_la <- sst_fall_la %>% group_by(rep_number) %>% do(model = do_CV(data = ., id = "native_id", group = "CV_grp_st_la", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_fall_CV_la <- st_fall_CV_la %>% rename(model_run = rep_number) #change column name so there isn't a conflict when I unnest it

st_fall_CV_la <- st_fall_CV_la %>%
  unnest(cols = c(model))

#---compare st_summer with gold-----------------------------------------#
#appending the predictions from the balanced short term campaign to the st_summer_CV df for comparison
gold_preds_la <- gold_CV_la %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds_la = cvpreds)  #select "model run" as well as native_id
st_fall_CV_la <- left_join(st_fall_CV_la, gold_preds_la, by = "native_id" ) 


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the balanced short term campaign to do method a

gold_obs_la <- gold_CV_la %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs_la = log_avg)
st_fall_CV_la <- left_join(st_fall_CV_la, gold_obs_la, by = "native_id" ) 
# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_fall_CV_la$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_fall_CV_la %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs_la, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_fall_MSE_la <- data.frame(
  model_run = unique(st_fall_CV_la$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_fall_MSE_la_long <- st_fall_MSE_la %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_fall_CV_la %>% dplyr::select(gold_preds_la, cvpreds) %>% range()


fall_scatter_la<-ggplot(data = st_fall_CV_la, aes(x = gold_preds_la, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "Fall Campaign \nvs Gold Standard Predictions", 
         caption = "1:1 line is dashed") +
    theme_bw()


#This needs some aesthetic adjustments: choose between dots or lines, pick a color scale that's helpful, get rid of the funky legend, but it's cool!!
    #looks a lot like the balanced predictions from the Blanco paper. 

#replicate blanco figure 3
fall_boxplot_la<- ggplot(st_fall_MSE_la_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Fall Comparison") +
  theme_bw()


```



```{r}

#------------seasonal comparison with gold standard in LA county-------------

#-----------------------------------WINTER--------------------------------------
#need to group the st_summer data by "campaign" before doing regression

#apply the save CV groups as used in the gold standard model
sst_winter_la <- st_winter_la %>%
  mutate(CV_grp_st_la = CV_grp_st_la) %>% ungroup()

#in sample-predictions
fitted_models = sst_winter_la %>% group_by(rep_number) %>% do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_winter_CV_la <- sst_winter_la %>% group_by(rep_number) %>% do(model = do_CV(data = ., id = "native_id", group = "CV_grp_st_la", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_winter_CV_la <- st_winter_CV_la %>% rename(model_run = rep_number) #change column name so there isn't a conflict when I unnest it

st_winter_CV_la <- st_winter_CV_la %>%
  unnest(cols = c(model))

#---compare st_summer with gold-----------------------------------------#
#appending the predictions from the balanced short term campaign to the st_summer_CV df for comparison
gold_preds_la <- gold_CV_la %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds_la = cvpreds)  #select "model run" as well as native_id
st_winter_CV_la <- left_join(st_winter_CV_la, gold_preds_la, by = "native_id" ) 


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the balanced short term campaign to do method a

gold_obs_la <- gold_CV_la %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs_la = log_avg)
st_winter_CV_la <- left_join(st_winter_CV_la, gold_obs_la, by = "native_id" ) 
# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_winter_CV_la$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_winter_CV_la %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs_la, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_winter_MSE_la <- data.frame(
  model_run = unique(st_winter_CV_la$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_winter_MSE_la_long <- st_winter_MSE_la %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_winter_CV_la %>% dplyr::select(gold_preds_la, cvpreds) %>% range()


winter_scatter_la<- ggplot(data = st_winter_CV_la, aes(x = gold_preds_la, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "Winter Campaign \nvs Gold Standard Predictions", 
         caption = "1:1 line is dashed") +
    theme_bw()


#This needs some aesthetic adjustments: choose between dots or lines, pick a color scale that's helpful, get rid of the funky legend, but it's cool!!
    #looks a lot like the balanced predictions from the Blanco paper. 

#replicate blanco figure 3
winter_boxplot_la<- ggplot(st_winter_MSE_la_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Winter Comparison") +
  theme_bw()


```




```{r}
#------------seasonal comparison with gold standard in LA county-------------

#-----------------------------------SPRING--------------------------------------
#need to group the st_summer data by "campaign" before doing regression

#apply the save CV groups as used in the gold standard model
sst_spring_la <- st_spring_la %>%
  mutate(CV_grp_st_la = CV_grp_st_la) %>% ungroup()

#in sample-predictions
fitted_models = sst_spring_la %>% group_by(rep_number) %>% do(model = lm(frml, data = .))

#cross-validated predictions treating each sample "campaign" separately
st_spring_CV_la <- sst_spring_la %>% group_by(rep_number) %>% do(model = do_CV(data = ., id = "native_id", group = "CV_grp_st_la", formula = frml))
#produces a df for each result

#next step is to restructure the result back into a single long_df

st_spring_CV_la <- st_spring_CV_la %>% rename(model_run = rep_number) #change column name so there isn't a conflict when I unnest it

st_spring_CV_la <- st_spring_CV_la %>%
  unnest(cols = c(model))

#---compare st_summer with gold-----------------------------------------#
#appending the predictions from the balanced short term campaign to the st_summer_CV df for comparison
gold_preds_la <- gold_CV_la %>%
  dplyr::select("native_id", "cvpreds") %>% rename(gold_preds_la = cvpreds)  #select "model run" as well as native_id
st_spring_CV_la <- left_join(st_spring_CV_la, gold_preds_la, by = "native_id" ) 


#calculating the RMSE and R2 two ways
  #a) comparing the campaign predictions to the observations from the gold standard 
  #b) comparing the predictions from each campaign to the observations from that campaign

#need to append the observations from the balanced short term campaign to do method a

gold_obs_la <- gold_CV_la %>%
  dplyr::select("native_id", "log_avg") %>% rename(gold_obs_la = log_avg)
st_spring_CV_la <- left_join(st_spring_CV_la, gold_obs_la, by = "native_id" ) 
# Initialize empty lists to store the results for both methods
results_a <- list()
results_b <- list()

# Loop through each unique model_run
for (run in unique(st_spring_CV_la$model_run)) {
  # Filter the data for the current model_run
  data_subset <- st_spring_CV_la %>% filter(model_run == run)
  
  # Calculate the MSE for method A
  mse_result_a <- get_MSE(data_subset$gold_obs_la, data_subset$cvpreds)
  results_a[[run]] <- mse_result_a
  
  # Calculate the MSE for method B
  mse_result_b <- get_MSE(data_subset$log_avg, data_subset$cvpreds)
  results_b[[run]] <- mse_result_b
}

# Convert the lists to a data frame with separate columns for each method
st_spring_MSE_la <- data.frame(
  model_run = unique(st_spring_CV_la$model_run),
  RMSE_a = sapply(results_a, `[`, 1),
  MSE.R2_a = sapply(results_a, `[`, 2),
  reg.R2_a = sapply(results_a, `[`, 3),
  RMSE_b = sapply(results_b, `[`, 1),
  MSE.R2_b = sapply(results_b, `[`, 2),
  reg.R2_b = sapply(results_b, `[`, 3)
)

#pivot longer for plotting
st_spring_MSE_la_long <- st_spring_MSE_la %>%
  pivot_longer(
    cols = c(RMSE_a, MSE.R2_a,reg.R2_a, RMSE_b, MSE.R2_b, reg.R2_b),
    names_to = c("metric", "method"),
    names_sep = "_",
    values_to = "value"
  ) 

#Plotting (replicate figure 1 in blanco paper)
# get range for plot
r <- st_spring_CV_la %>% dplyr::select(gold_preds_la, cvpreds) %>% range()


spring_scatter_la<-ggplot(data = st_spring_MSE_la_long, aes(x = gold_preds_la, y = cvpreds, color = as.factor(model_run))) +
    geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) +
    lims(x= r, y = r) +
    coord_fixed() +
    geom_smooth(method = 'lm', se = F, alpha = 0.5, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "darkgray", linetype = "dashed") +
    labs(title = "Spring Campaign \nvs Gold Standard Predictions",
         caption = "1:1 line is dashed") +
    theme_bw()


#This needs some aesthetic adjustments: choose between dots or lines, pick a color scale that's helpful, get rid of the funky legend, but it's cool!!
    #looks a lot like the balanced predictions from the Blanco paper. 

#replicate blanco figure 3
spring_boxplot_la<-ggplot(st_spring_MSE_la_long, aes(x = method, y = value, color = method)) +
  geom_boxplot() +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[1]], metric = "RMSE"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  geom_point(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[2]], metric = "MSE.R2"), aes(x = method, y = value, color = method), size = 3, shape = 5) +
  facet_wrap(~metric) +
  labs(title = "Spring Comparison") +
  theme_bw()
```





```{r, echo = F, include = T, fig.height = 7, fig.width = 7, warning = F, message = F}
#LA-county scatter predictions
#trying to make a better plot

#add a season column to each one
st_spring_CV_la <- st_spring_CV_la %>% mutate(season = "Spring")
st_summer_CV_la <- st_summer_CV_la %>% mutate(season = "Summer")
st_fall_CV_la <- st_fall_CV_la %>% mutate(season = "Fall")
st_winter_CV_la <- st_winter_CV_la %>% mutate(season = "Winter")
st_la_CV <- st_la_CV %>% mutate(season = "Balanced")

#bind the df into one
st_all_CV_la <- rbind(st_la_CV, st_spring_CV_la, st_summer_CV_la, st_fall_CV_la, st_winter_CV_la)

#factor to control the order
st_all_CV_la <- st_all_CV_la %>%
  mutate(season = factor(season, levels = c("Spring", 
                                            "Summer", 
                                            "Fall", 
                                            "Winter", 
                                            "Balanced")))


all_scatter_la <-ggplot(data = st_all_CV_la, aes(x = gold_preds_la, y = cvpreds, color = as.factor(model_run))) +
    #geom_point(shape = "o", alpha = 0.7, show.legend = FALSE) + #commented out the points to make the plot a little cleaner. 
    lims(x= r, y = r) +
    geom_smooth(method = 'lm', se = F, alpha = 0.4, show.legend = FALSE, linewidth = 0.3) +
    facet_wrap(~season, scales = "free") + 
    geom_abline(linetype = "dashed", slope = 1, intercept = 0, color = "darkgray") +
    labs(title = "Fig. X: Best fit lines of cross-validated short-term predictions for 30 campaigns\nvs. gold standard predictions for ln(NOx)", 
         subtitle = "Restricted to LA County",
         x = "Gold standard predicted ln(NOx) (ln(ppb))",
         y = "Campaign Predicted ln(NOx) (ln(ppb))",
         caption = "1:1 line is dashed") +
    theme_bw()

all_scatter_la

#definitely a more heterogeneity in slope here (one differene is LOO CV)

```

```{r, echo = F, include = T, fig.height = 7, fig.width = 7}

#bind the MSE df together for a combined boxplot
#add a season column to each one
st_spring_MSE_la_long <- st_spring_MSE_la_long %>% mutate(season = "Spring")
st_summer_MSE_la_long <- st_summer_MSE_la_long %>% mutate(season = "Summer")
st_fall_MSE_la_long <- st_fall_MSE_la_long %>% mutate(season = "Fall")
st_winter_MSE_la_long <- st_winter_MSE_la_long %>% mutate(season = "Winter")
st_la_MSE_long <- st_la_MSE_long %>% mutate(season = "Balanced")

st_all_MSE_la_long <- rbind(st_spring_MSE_la_long, st_summer_MSE_la_long, st_fall_MSE_la_long, st_winter_MSE_la_long, st_la_MSE_long)

#factor to control the order
st_all_MSE_la_long <- st_all_MSE_la_long %>%
  mutate(season = factor(season, levels = c("Spring", 
                                            "Summer", 
                                            "Fall", 
                                            "Winter", 
                                            "Balanced")))

#replicate blanco figure 3
st_all_boxplot_la <- ggplot(st_all_MSE_la_long, aes(x = season, y = value, color = season)) +
  geom_boxplot() +
  geom_hline(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[1]], metric = "RMSE"), aes(yintercept = value, linetype = "dotted"), show.legend = T) +
  geom_hline(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[2]], metric = "MSE.R2"), aes(yintercept = value, linetype = "dashed"), show.legend = T) +
  geom_hline(data = data.frame(method = c("a", "b"), value = gold_MSE_la[[3]], metric = "reg.R2"), aes(yintercept = value, linetype = "dotdash"), show.legend = T) +
  facet_grid(metric~method, scales = "free") +
  labs(title = "RMSE and R2 Comparison - Los Angeles County",
       caption = "Method A = Short-term campaign predictions compared to gold-standard obs\nMethod B = Short-term campaign predictions compared to same campaign observations\nHorizontal lines show the gold-standard performance", 
       x = "Short-term Campaign",
       y = "Value",
       color = "Season") +
  scale_linetype_manual(name='',
                      labels = c("Gold standard RMSE", "Gold Standard MSE-R2", "Gold Standard Reg. R2"), 
                      values=c("dotted", "dashed", "dotdash")) +
  theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()# Remove x-axis text labels
  )

st_all_boxplot_la



```



Figures X and X both show the restricted LA county models' performances. In figure X, the box plots, we see the same trend we saw with the statewide data sets. The summer and spring models describe nearly zero percent of the variability in the data set (looking at the pink R squared), and the fall and winter models predict some of the variability, but much less than the gold standard model.

Figure X, with scatter plots, shows the same trend that we saw in the statewide data with under- and over- predictions. The summer and spring models under-predict ln(NOX) concentrations, again, while the winter and fall models over-predict. In good news, the short term balanced model continues to agree with the gold standard model and offer comparable R-squared and RMSE statistics. 




```{r prediction.maps, echo = F, include = F}

#this part is just a little messy because I'm working in a seperate file. Will be able to do it more cleanly in original doc, but I want to keep the same coord system I've already got working here. 

#add the season name back into each seasonal CV df

st_spring_CV_la <- st_spring_CV_la %>% mutate(season = "Spring")
st_summer_CV_la <- st_summer_CV_la %>% mutate(season = "Summer")
st_fall_CV_la <- st_fall_CV_la %>% mutate(season = "Fall")
st_winter_CV_la <- st_winter_CV_la %>% mutate(season = "Winter")
st_la_CV <- st_la_CV %>% mutate(season = "Balanced")
gold_CV_la <- gold_CV_la %>% mutate(season = "Gold")

#bind the st_ df together
la_all_campaign <- rbind(st_la_CV, st_spring_CV_la, st_summer_CV_la, st_fall_CV_la, st_winter_CV_la)



#generate and average value per across model runs for the st campaigns and remove unncessary columns
la_all_campaign <- la_all_campaign %>% group_by(native_id, season) %>%
  mutate(
    avg_conc = mean(annual_avg),
    log_conc =log(annual_avg),
    cvpred = mean(cvpreds)
  ) %>%
  ungroup() %>%
  distinct(native_id, season, .keep_all = TRUE) %>% dplyr::select(-model_run, -annual_avg, -log_avg, -seed, -rep_number, -gold_preds_la, -gold_obs_la, -cvpreds, -CV_grp_st_la)

#rename and reorder columns from gold so it matches the st. 
gold <- gold_CV_la %>% rename(avg_conc = annual_avg, log_conc = log_avg, cvpred = cvpreds) %>% dplyr::select(-CV_grp_la)
# Reorder df2 to match the column order of df1
gold <- gold[, colnames(la_all_campaign)]

# Check the result
head(gold)
head(la_all_campaign)

#add the gold to the df
la_all_campaign <- rbind(la_all_campaign, gold)

labert_only <- geom %>% dplyr::select(native_id, lambert_x, lambert_y)

#join to the geometry from the the sf
la_all_campaign_sf <- left_join(la_all_campaign, labert_only, by = "native_id")
la_all_campaign_sf <- st_as_sf(la_all_campaign_sf, coords=c('lambert_x','lambert_y'), crs=lambert_proj) %>% 
  dplyr::select(-geometry.x, -geometry.y)

#leaving in lambert (m) for kriging

```

```{r setup.grid, echo = F, include = F}
#need to change some variable names so that they match our model covariates 
la_grid <- la_grid %>% rename(pop_s05000 = Pop_5000, m_to_coast = D2C, m_to_a1 = D2A1) 

#need to transform covars the same as mercer (or untransform mercer)

#attempting to untransform based on the descriptions of the functional variables in Mercer
la_grid$pop_s05000 <- la_grid$pop_s05000 * 100000 #pop was 1/100000
la_grid$m_to_coast <- la_grid$m_to_coast * 100000 #dist. in m was 1/100000
la_grid$m_to_a1 <- 10^(la_grid$m_to_a1) #value was log10 transformed



# Download the land polygon data as an sf multipolygon
# the CRS for this is in lat/long degrees
land <- ne_download(scale = "large", type = "land", category = "physical", returnclass = "sf")

# Crop the land area to the bounding box of the LA grid to reduce processing time
# have to convert the la_grid to the same degrees as LA
land <- suppressWarnings(st_crop(land, st_bbox(la_grid)))  

# Visualize cropped land area (optional)
ggplot(land) + geom_sf()

# Filter la_grid to keep only points that intersect with land
la_grid <- la_grid[st_within(la_grid, land) %>% lengths() > 0,]

# Visualize grid land locations (zoom in)
ggplot(la_grid) + geom_sf(size=0.001)

# ---- LA Map Setup ----
# Define a bounding box (min & max X and Y) with a 10,000m buffer around `la_grid`
map_bbox <- la_grid %>%
  # convert from degrees to meters
  st_transform(crs = lambert_proj) %>%
  # add a buffer around the area for visualization purposes
  st_buffer(dist = 1000) %>%
  # convert back to original CRS
  st_transform(crs = latlong_proj) %>%
  # take the min/max X/Y
  st_bbox()

map_bbox

# Base map setup with ggplot2 and ggspatial using OSM tiles
g <- ggplot() +
  ggspatial::annotation_map_tile(type = "osm", zoom = 10) +
  labs(title = "LA Grid with Map") +
  theme_minimal()

# # alternative background map with the maptiles package
# tiles <- maptiles::get_tiles(x = st_bbox(la_grid), provider = "OpenStreetMap")
# g <- ggplot() +
#   # Add basemap tiles as background
#   layer_spatial(tiles)

# Plot background map and the LA grid (zoom in)
g + 
  geom_sf(data = la_grid, size=0.001)


# Add NOx data with additional map elements

testing_sf <- annual_sf %>% filter(County == "Los Angeles") %>% st_drop_geometry()
testing_sf <- st_as_sf(testing_sf, coords=c('lambert_x','lambert_y'), crs=lambert_proj)
testing_sf <- st_transform(testing_sf, crs = latlong_proj)

class(testing_sf)

g + 
  geom_sf(data = testing_sf, aes(color = annual_avg)) + 
  scale_color_viridis_c() +  # Color-friendly scale
  theme_void() +  # Clean layout for map aesthetics
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 1)
  ) + 
  ggspatial::annotation_scale(location = "bl", width_hint = 0.3, unit_category = "imperial") +  # Scale in miles
  ggspatial::annotation_north_arrow(location = "tr", which_north = "true") +  # North arrow
  labs(title = "Map of Los Angeles with the Seasonal Data",
       col="NOx (ppb)"
       )

#looking at monitors that fit on the grid

g + geom_sf(data = la_grid, size=0.001) +
  geom_sf(data = testing_sf, color = "green")

#9 of the 12 monitors fit on the grid. 3 of the LA monitors are out of bounds.

#just want to check if there are any other that might fit on the map in the rest of socal

testing2_sf <- annual_sf %>% filter(County %in% counties) %>% st_drop_geometry()
testing2_sf <- st_as_sf(testing2_sf, coords=c('lambert_x','lambert_y'), crs=lambert_proj)
testing2_sf <- st_transform(testing2_sf, crs = latlong_proj)

g + geom_sf(data = la_grid, size=0.001) +
  geom_sf(data = testing2_sf, color = "green")

#There are more than 9 monitors on the grid
#looks like there are 13 fully on the grid and just on the edge. 

#Finding those monitors

gridstrict_bbox <- st_bbox(la_grid)
grid_monitors <- st_crop(testing2_sf, gridstrict_bbox)
#9 in LA county, 3 in orange co, and 1 in san bernardino co. 

#it would take a little effort, but not that much to do the kriging with these 13 monitors instead of the LA Co 12, which actually ends up only being 9 in the grid. 

#will test just with the gold - can replicate later with others if we think it's worthwhile. If it looks better it won't be that much to add the other seasons depending on what we want to do

gold_la_monitors <- right_join(long_term_gold, grid_monitors,  by = "native_id") %>% dplyr::select(-annual_avg.y, - County.y, - geometry.x) %>% rename(geometry = geometry.y, avg_conc = annual_avg.x, log_conc = log_avg, County = County.x)

st_crs(gold_la_monitors) 

gold_la_monitors <- st_as_sf(gold_la_monitors, crs = latlong_proj)

#df of just monitors to add as points
monitors_df <- ca_covariates %>% dplyr::select(native_id, latitude, longitude) %>%
  st_drop_geometry()

 
```


```{r def.mapping.functions, echo = F, include = F}
#---this section tests out maping predictions using lm and then defintes functions to make it easy to do repeatedly for the different data sets. ---#

#--------- MAIN USE------------#
#   map_data(model)  => function that takes an lm model and returns a df that can be mapped
#   plot_map(data)  => function that takes the output from map_data() and then makes a map


#include = F because there are some test maps in the this chunk that don't need to be rendered. 

#trying regular prediction mapping

#fit the model
#lm_gold_standard <- lm(frml, data = long_term_gold)) #commented out because we already have the model

la_grid <-st_transform(la_grid, crs = latlong_proj) #change to lat long proj

#predict for new data
la_grid_gold_pred <- la_grid %>% mutate(preds = predict(lm_gold_standard, la_grid))

la_grid_gold_pred_df <- as.data.frame(st_coordinates(la_grid_gold_pred))
la_grid_gold_pred_df$lnNOx <- la_grid_gold_pred$preds 
la_grid_gold_pred_df$NOx <- exp(la_grid_gold_pred$preds)

#function to make pretting the data easy
map_data <- function(model) { #specify the lm model to use for predicting on la_grid
  pred_data <- la_grid %>% mutate(preds = predict(model, la_grid)) # generate predictions
  pred_data_df <- as.data.frame(st_coordinates(pred_data)) #convert to data frame for plotting
  pred_data_df$lnNOx <- pred_data$preds #add the predictions (ln scale)
  pred_data_df$NOx <- exp(pred_data$preds) #convert pred to native scale
  
  return(pred_data_df) #return a df wtih X, Y, lnNOx and NOx
}

#test the function

test_df <- map_data(lm_gold_standard) #yay!!

gold_predict <- g + 
  # Set map extent and CRS (bounding box error otherwise results in no background map) 
  coord_sf(xlim = c(map_bbox["xmin"], map_bbox["xmax"]), 
           ylim = c(map_bbox["ymin"], map_bbox["ymax"]), 
           crs = 4326) +  
  geom_tile(data = test_df, aes(x = X, y = Y, fill = lnNOx), 
            alpha=0.2,
            width = 0.01, height = 0.01 # Adjust width and height as needed
            ) +  
  geom_point(data = monitors_df, aes(x = longitude, y = latitude), size = 2, shape = 18, color = "black") + #add the monitors as points just to see. 
  # color friendly color scale
  scale_fill_viridis_c(option = "plasma") + 
  labs(title = "Map of Los Angeles with Predictions - Gold Standard",
       col="ln(NOx(ppb))",
       caption = "Diamonds indicate monitor locations",
       x = "",
       y = ""
       ) +
  theme_minimal()

gold_predict

#function that will plot the map using prepared data from the map_data() function
#requires a bounding box be already set called map_bbox
#plots predictions on the native scale
plot_map <- function(data) { #specify the data to use for the map
  
  g <- ggplot() +
  ggspatial::annotation_map_tile(type = "osm", zoom = 10) +
  labs(title = "LA Grid with Map") +
  theme_minimal()
  
  plot <- g + # Set map extent and CRS (bounding box error otherwise results in no background map) 
  coord_sf(xlim = c(map_bbox["xmin"], map_bbox["xmax"]), 
           ylim = c(map_bbox["ymin"], map_bbox["ymax"]), 
           crs = 4326) +  
  geom_tile(data = data, aes(x = X, y = Y, fill = NOx), #plot on native scale
            alpha=0.2,
            width = 0.01, height = 0.01 # Adjust width and height as needed
            ) +  
  geom_point(data = monitors_df, aes(x = longitude, y = latitude), size = 2, shape = 18, color = "black") + #add the monitors as points just to see. 
  # color friendly color scale
  scale_fill_viridis_c(option = "plasma") + 
  labs(title = "Map of Los Angeles with Predicted NOx",
       fill ="NOx (ppb)",
       caption = "Diamonds indicate monitor locations",
       x = "",
       y = ""
       ) +
  theme_minimal() +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
  
  return(plot) 
}

test_map <- plot_map(test_df)

test_map + labs(title = "Map of Los Angeles with Predictions - Gold")

```


```{r generate.maps, echo = F, include = F}

#testing using full state models

gold_map <- plot_map(map_data(lm_gold_standard))
balanced_map <- plot_map(map_data(lm_st_balanced))
spring_map <- plot_map(map_data(lm_st_spring))
summer_map <- plot_map(map_data(lm_st_summer))
fall_map <- plot_map(map_data(lm_st_fall))
winter_map <- plot_map(map_data(lm_st_winter))

gold_map + labs(title = "Map of Los Angeles with Predictions - Gold")
balanced_map + labs(title = "Map of Los Angeles with Predictions - Balanced")
spring_map + labs(title = "Map of Los Angeles with Predictions - Spring")
summer_map + labs(title = "Map of Los Angeles with Predictions - Summer")
fall_map + labs(title = "Map of Los Angeles with Predictions - Fall")
winter_map + labs(title = "Map of Los Angeles with Predictions - Winter")

#They all look pretty much the same and aren't interesting. The LA specific ones are much more interesting!


```


```{r maps.la.model, echo = F, include = T}
#repeat using the LA specific models

la_gold_map <- plot_map(map_data(lm_gold_la))
la_balanced_map <- plot_map(map_data(lm_st_la))
la_spring_map <- plot_map(map_data(lm_st_spring_la))
la_summer_map <- plot_map(map_data(lm_st_summer_la))
la_fall_map <- plot_map(map_data(lm_st_fall_la))
la_winter_map <- plot_map(map_data(lm_st_winter_la))

la_gold_map + labs(title = "Map of Los Angeles with NOx Predictions - Gold Standard")
la_balanced_map + labs(title = "Map of Los Angeles with NOx Predictions - Balanced ")
la_spring_map + labs(title = "Map of Los Angeles with NOx Predictions - Spring")
la_summer_map + labs(title = "Map of Los Angeles with NOx Predictions - Summer")
la_fall_map + labs(title = "Map of Los Angeles with NOx Predictions - Fall")
la_winter_map + labs(title = "Map of Los Angeles with NOx Predictions - Winter")


#so interesting!

#I think the faceted plots with ln nox are better for the results, but these might go in the appendix, because they are just cool and really show what is going on. 
```
The color ramps arent the same size. Summer is lower over all, but the differences are really seen a long the roads. Winter is the highest (the blue in winter is higher than the yellow in summer), but there is less variation and a peak near downtown.
Spring and fall are midway in-between with spring being more like summer and fall being more like winter.
Fall looks the closest to balanced

```{r, fig.height = 7, fig.width = 7, echo = F, include = T}

#trying one more thing to see if it makes the color ramps easier to compare

#generate df only and add the season variable 
la_gold_map_df <- map_data(lm_gold_la) %>% mutate(season = "Gold")
la_balanced_map_df <- map_data(lm_st_la) %>% mutate(season = "Balanced")
la_spring_map_df <- map_data(lm_st_spring_la) %>% mutate(season = "Spring")
la_summer_map_df <- map_data(lm_st_summer_la) %>% mutate(season = "Summer")
la_fall_map_df <- map_data(lm_st_fall_la) %>% mutate(season = "Fall")
la_winter_map_df <- map_data(lm_st_winter_la) %>% mutate(season = "Winter")

#bind into one df
la_all_map_df <- rbind(la_gold_map_df,
                       la_balanced_map_df,
                       la_spring_map_df,
                       la_summer_map_df,
                       la_fall_map_df,
                       la_winter_map_df
                       )

#create factor variable to control the order in faceting

la_all_map_df <- la_all_map_df %>%
  mutate(season = factor(season, levels = c("Spring", 
                                            "Summer", 
                                            "Fall", 
                                            "Winter", 
                                            "Balanced", 
                                            "Gold")))

#make the map using faceting

nox_map <- g + # Set map extent and CRS (bounding box error otherwise results in no background map) 
  coord_sf(xlim = c(map_bbox["xmin"], map_bbox["xmax"]), 
           ylim = c(map_bbox["ymin"], map_bbox["ymax"]), 
           crs = 4326) +  
  geom_tile(data = la_all_map_df, aes(x = X, y = Y, fill = NOx), #plot on native scale
            alpha=0.2,
            width = 0.01, height = 0.01 # Adjust width and height as needed
            ) +  
  geom_point(data = monitors_df, aes(x = longitude, y = latitude), size = 2, shape = 18, color = "black") + #add the monitors as points just to see. 
  # color friendly color scale
  scale_fill_viridis_c(option = "plasma") + 
  facet_wrap(~season) + 
  labs(title = "Map of Los Angeles with Predicted NOx by Season",
       fill ="NOx (ppb)",
       caption = "Diamonds indicate monitor locations",
       x = "",
       y = ""
       ) +
  theme_void() +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

#nox_map

#plotting on the same scale squashes variability, because the color ramp isn't long enough. 

#trying the same on ln scale to see if it enhances variability while keeping the scale continuous

lnnox_map <- g + # Set map extent and CRS (bounding box error otherwise results in no background map) 
  coord_sf(xlim = c(map_bbox["xmin"], map_bbox["xmax"]), 
           ylim = c(map_bbox["ymin"], map_bbox["ymax"]), 
           crs = 4326) +  
  geom_tile(data = la_all_map_df, aes(x = X, y = Y, fill = lnNOx), #plot on native scale
            alpha=0.2,
            width = 0.01, height = 0.01 # Adjust width and height as needed
            ) +  
  geom_point(data = monitors_df, aes(x = longitude, y = latitude), size = 2, shape = 18, color = "black") + #add the monitors as points just to see. 
  # color friendly color scale
  scale_fill_viridis_c(option = "plasma") + 
  facet_wrap(~season, nrow = 3) +
  labs(title = "Map of Los Angeles with Predicted ln(NOx) by Season",
       fill ="ln (NOx (ppb))",
       caption = "Diamonds indicate monitor locations",
       x = "",
       y = ""
       ) +
  theme_void() +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

lnnox_map #like this one a lot. They do look cool on the native scale, but I think this is the best way to show them together and still get the variation without the confusion of different color ramps. 

#this is better and helpful for seeing relative comparison, but I really like th distinctnes on the native scale when the scales aren't set. But is it misleading to have different color scales?

```

# Discussion and Conclusion

**Statewide Short-Term Dataset Performance**

Our short-term, temporally-balanced dataset produced stronger predictions than any of the season-specific datasets, as indicated by its lower RMSE and higher R^2 values. Approximately 21% of the variation in the NOx predictions was explained by the model predictors in the temporally-balanced dataset and, on average, the predicted NOx concentrations were 0.55 log-ppb away from the observed values. When compared against the gold-standard dataset, the short-term temporally-balanced dataset performed reasonably well but showed small reductions in R^2 (0.26 vs. 0.21) and increases in RMSE (0.49 vs. 0.55 log-ppb). 

In season-specific datasets, little to none of the variation in cross-validated NOx predictions was explained by the model predictors, with R^2 values generally hovering around 0.0 (summer, fall, and spring) or 0.1 (winter) when compared against the gold standard dataset. Short-term monitoring campaigns using only summer or spring data consistently under-estimated the gold standard annual average concentrations, whereas short-term campaigns using only winter or fall data consistently over-estimated the gold standard annual average concentrations. 

Taken together, these data suggest that a seasonally-balanced monitoring campaign will predict annual NOx concentrations with improved accuracy compared to a monitoring campaign with data collected in just one season. These results are consistent with prior data demonstrating that NOx concentrations are typically highest in the winter and lowest in the summer and spring (Lange et al. 2022; EPA 2023). They are also consistent with the findings of Blanco et al. (2023), who reported that a temporally-balanced sampling design was important in accurately predicting annual average NOx concentrations during mobile monitoring campaigns.

Notably, all models, including those using the gold standard dataset, had low R^2 values, indicating that just a small proportion of variability in NOx concentrations were explained by model covariates. This could be related to our model selection strategy, which is discussed in greater detail in subsequent sections. Importantly, Blanco et al. (2023) reported improved coefficients of determination compared to the present analysis, though the values were still somewhat low and tended to hover around approximately 0.35 to 0.45. In Blanco et al. (2023), authors included hundreds of geographic covariates in their model related to land-use, road proximity, and population density. The inclusion of a robust set of geographic covariates likely contributed to the improved performance of models in Blanco et al. (2023) versus the present analysis, though low coefficients of determination in both analyses suggest that the increased number of geographic covariates alone do not fully capture the variability in NOx concentrations. It is possible that other temporal or spatial factors uncaptured by the covariates in either analysis are additionally contributing to the observed variability in NOx concentrations. 

**Los Angeles County Short-Term Dataset Performance** 

Similar trends were observed for our Los Angeles-specific analysis when compared to the statewide analyses. Season-specific datasets continued to under- or over-predict NOx concentrations, with little to none of the variability in gold standard NOx concentrations described by the models with season-specific data. Notably, the Los Angeles analyses included annual aggregate estimates from just 12 sensors. The decreased sample size in this analysis increased the variability of the data compared to the statewide analysis. Additionally, our use of just 12 sensors to represent concentrations across all of Los Angeles county was likely not adequate in capture the true spatial variability in the data. Los Angeles county has a population of more than 9.5 million people across a vast geographic area (U.S. Census 2023), therefore more complex pollution heterogeneity is expected than what could reasonably be captured by 12 sensors. 

**Model Selection** 

We used forward stepwise regression to select model covariates from among the seven used by Mercer et al. (2011) in their common model describing all seasons. We chose to use stepwise regression to select covariates from among the common model covariates rather than incorporating all seven used by Mercer et al. (2011) due to the limited data available in our LA county-specific analysis, which included just 12 sensors. The Mercer et al. (2011) common model covariates were used because of their efficacy in predicting seasonal NOx concentrations in California, as determined through a robust selection process that determined which covariates performed well across all seasons. Although the Mercer et al. (2011) covariates performed reasonably well in the Mercer et al. (2011) analysis, similar covariates in our model explained just small proportion of the variability in NOx predictions, as demonstrated by our small-to-moderate coefficients of determination generated for cross-validated predictions across all seasons. It is likely that some of the covariates not selected for inclusion into our model may have explained a greater portion of the variability in NOx concentrations; this is particularly important to consider, given that our primary analysis focused on the entire state of California, whereas the Mercer et al. (2011) analysis evaluated only Los Angeles. The geographic differences, including potential differences in spatial correlation, could therefore contribute to the varied performance of the covariates across analyses. 

Our use of forward selection stepwise regression for covariate selection was beneficial in providing information on the bias-variance-tradeoff as covariates were systematically incorporated into the model and assessed. Through this process, we were able to objectively select three of the seven Mercer et al. (2011) covariates based on their contributions to model accuracy and variability. Although forward selection was a reasonable choice for model selection based on our goal (i.e. narrowing down covariates from those in the Mercer et al. common model) and the statistical tools available to us, it is possible that another covariate selection method would have provided a stronger prediction model in the present analysis. For example, in Blanco et al. (2023), authors used partial least squares (PLS) regression models to summarize hundreds of geographic covariates, rather than a priori selecting a handful of covariates based on previous findings published in the literature. The use of PLS across hundreds of covariates enhanced the ability of the model to capture variability in NOx concentrations and likely accounted for their improved prediction performance when compared against the present analysis. 

**Comparison of Findings to Literature** 

Blanco et al (2023) analyzed their short-term campaigns versus a gold standard using partial least squares (PLS), and had similar but slightly better results than our short term balanced and gold standard models. Blanco et als (2023) gold standard model had a MSE-based R squared of 0.46 and a RMSE of 7.2 ppb. The balanced, short-term model had the best R squared and RMSE out of all of the short-term campaigns, as well (Blanco et al. 2023). Although our gold standard model did not perform quite as well as the Blanco et al. (2023) PLS gold standard model, we believe our results support the same conclusions with the same data. 


**Strengths and Limitations** 

A limitation of our analysis is that our dataset was not spatially rich, therefore we were limited in where we could predict and which counties we could evaluate in our subanalyses. We did not have enough sensors in counties other than LA county to compare predictions. The low numbers of sensors in our LA county models could have limited our sensitivity analysis, and may have limited our confidence in the comparison between the LA county short term balanced model and the LA county gold standard model. Our Gold standard model in Los Angeles had just 12 data points for annual aggregate concentration, so it may be misleading to consider its predictions the most well-founded predictions compared to the short term balanced model. 

Additionally, as noted earlier, our model selection method was not complex; we simply used the previously selected model from Mercer et al (2011) paired with forward, stepwise model selection to further reduce the number of covariates included in our final common model. The subset of our land use covariates to the eight reflecting the Mercer et al (2011) common model may have erroneously eliminated covariates that would have explained more of the variation between monitor locations. A more in-depth model selection approach may have found a better fit to the models.

Finally, when creating our season-specific datasets, we created seasonal strata by dividing the weeks of the year into four segments based on their numerical order, each with 13 weeks. Though this provided us with a well-balanced and unbiased division of the dataset, there was no distinct scientific or meteorological significance to these cutpoints. In future analyses, we may consider defining season using specific calendrical definitions, looking at cutpoints based on temperature or precipation changes, or based on social behaviors and patterns (e.g. defining summer as when school is out). Although our definition of season was based solely on an even division of the data in temporal order, we feel confident that seasonal trends were reasonably well captured in our data subsets because our data generally followed the seasonal NOx trends observed in prior published data (i.e. lower summer concentrations and higher winter concentrations). 

**Broader Implications and Future Research** 

Our findings are important in informing the sampling design for future monitoring campaigns. Our results suggest that future monitoring campaigns must include sampling balanced across all seasons, rather than overrepresenting or restricting to sampling within a single season. This finding provides important insight for future investigators, who will need to incorporate sampling over the course of a year rather than within a more constrained timeframe. Additionally, our findings are valuable in reaffirming that a temporally-balanced dataset can adequately predict long-term annual averages; this reinforces prior findings that suppoted the use of short-term mobile monitoring campaigns to characterize gold standard concentration measurements.

We recommend future research that compares the short-term monitoring campaigns across counties in California. This analysis would be an interesting next step to further compare how the short term balanced and seasonal datasets perform when restricted to different geographic regions in California; this could be especially relevant for counties that have lower annual variability in meteorological parameters. However, a different data set may be necessary for this analysis, as the present dataset was limited in the total number of county-specific sensors outside of Los Angeles.

**Conclusion** 

Overall, our findings re-iterate the findings of Blanco et al, 2023, whose methods inspired many of our analysis steps. We see seasonal variability in NOx concentrations at the statewide and LA county level, which leads to the importance of a sampling campaign which samples throughout all seasons. Both at the statewide and LA county level, we see reduced performance of our sampling campaign design when it was restricted to only one season. Our balanced short term sampling campaign performed comparably to our gold-standard model on a statewide and county level, confirming that a sampling campaign can collect a smaller number of samples on which to predict NOx concentrations, so long as the sampling design is temporally balanced. 


# Author Contribution Statement

**Core Contributions** 

Analysis: 

*Callan:* Met with professors at office hours to discuss analytical approach; attended all group meetings to discuss analytical approach; drafted first iteration of day-of-week  vs. weekday/weekend predictions with cross-validation and ANOVA (not incorporated: later revised plan); conducted sensitivity analysis for Los Angeles county, including creation of short- and long-term temporally balanced datasets specific to Los Angeles, summarizing short- and long-term temporally balanced datasets within LA, generating LA-specific NOx predictions and taking the first pass at cross-validation, including making a new cross-validation group; calculated R2 and RMSE values for LA-specific NOx predictions

*Katie:* Met with professors at office hours and attended all group meetings. CA NOx monitor inclusion criteria validation. Wrote functions for random campaign sampling and data aggregation and summary statistics. Wrote template code for iterating CV function across campaigns and extracting the performance statistics. Figure design and all maps. 

*Abbie:* Met with professors at office hours and attended all group meetings. Set up the Github Project and Repository for the project; Implemented the forward stepwise selection of our model terms. Created the statewide seasonal data sets and models, including the comparisons with the gold standard data sets; created the LA county seasonal data sets and models, including the comparisons with the gold standard data sets; adjustment of our cross validation method for LA county, implementing a leave-one-out method instead of a 10 fold method.

Writing: 

*Callan:* Drafted and revised analysis plans (3 in total); wrote and revised methods section; wrote and revised introduction section; wrote portions of results section; wrote portions of discussion section

*Katie:* Added to the results, methods, and discussion, especially the portion about maps. Proof reading and editing for consistent interpretation between writing and code output. 


*Abbie:* Added to the results, especially with regards to the model selection and selection of model covariates, and seasonal models. Added to the discussion section with overall conclusions, limitations, future research suggestions, and current literature. final editing of plots and graphs. Review of citations.


**Additional Contributions**

*Callan:* reviewed literature for introduction section; reviewed and critiqued content; final editing and proofreading

*Katie:*

*Abbie:*


# Works Cited 

Blanco, M. N., Gassett, A., Gould, T., Doubleday, A., Slager, D. L., Austin, E., ... & Sheppard, L. (2022). Characterization of annual average traffic-related air pollution concentrations in the Greater Seattle Area from a year-long mobile monitoring campaign. Environmental science & technology, 56(16), 11460-11472.

Blanco, M.N., Doubleday, A., Austin, E., Julian D. Marshall, Edmund Seto, Timothy V., Larson & Lianne Sheppard  Design and evaluation of short-term monitoring campaigns for long-term air pollution exposure assessment. J Expo Sci Environ Epidemiol 33, 465473 (2023). https://doi.org/10.1038/s41370-022-00470-5.

Colorado Department of Public Health & Environment. 2024. Air toxics: monitoring. https://cdphe.colorado.gov/air-toxics/monitoring#:~:text=Mobile%20monitoring%20can%20capture%20%E2%80%9Cspatial,wind%20compared%20to%20stationary%20monitoring Accessed 12/3/2024.

Environmental Protection Agency. 2023. Overview of Nitrogen Dioxide (NO2) Air Quality in the United States. https://www.epa.gov/system/files/documents/2023-06/NO2_2022.pdf 

Lange, K., Richter, A., & Burrows, J. P. (2022). Variability of nitrogen oxide emission fluxes and lifetimes estimated from Sentinel-5P TROPOMI observations. Atmospheric Chemistry and Physics, 22(4), 2745-2767.

Mercer, L. D., Szpiro, A. A., Sheppard, L., Lindstrm, J., Adar, S. D., Allen, R. W., ... & Kaufman, J. D. (2011). Comparing universal kriging and land-use regression for predicting concentrations of gaseous oxides of nitrogen (NOx) for the Multi-Ethnic Study of Atherosclerosis and Air Pollution (MESA Air). Atmospheric Environment, 45(26), 4412-4420.
https://pmc.ncbi.nlm.nih.gov/articles/PMC3146303/pdf/nihms299256.pdf

United States Census. 2023. QuickFacts Los Angeles County, California. https://www.census.gov/quickfacts/fact/table/losangelescountycalifornia,losangelescitycalifornia,CA/PST045223 Accessed 11/30/2024.



## Appendix 

#---------------------------Reference Code-------------------------------------#


Sensors Kept in the Data after applying selection Criteria from Blanco et al (2022).
```{r, echo = FALSE}
kable(sensor_summary,
      col.names = c("Sensor ID", "Count", "Percent of Year with Data", "Percent of Data Positive", "Maximum Data Gap (days)"),
      digits = 2,
      caption = "Summary of Sensor Data for SEnsors Meeting Selection Criteria") %>% kable_styling()

```

Statewide Datasets - Gold Standard, Short term balanced, and seasonals
```{r}
#summary statistics
lt_gold_sum <- long_term_gold %>% summarise(min = min(annual_avg), 
                                            Q25 = quantile(annual_avg, probs = 0.25), 
                                            Q50 = quantile(annual_avg, probs = 0.5),
                                            Q75 = quantile(annual_avg, probs = 0.75),
                                            max = max(annual_avg),
                                            mean = mean(annual_avg),
                                            SD = sd(annual_avg)
                                            )
#lt_gold_sum
      #This matches the number reported in Blanco supplement S5

#summary stats for st_balanced
st_balanced_sum <- st_balanced %>% summarise(min = min(annual_avg), 
                                            Q25 = quantile(annual_avg, probs = 0.25), 
                                            Q50 = quantile(annual_avg, probs = 0.5),
                                            Q75 = quantile(annual_avg, probs = 0.75),
                                            max = max(annual_avg),
                                            mean = mean(annual_avg),
                                            SD = sd(annual_avg)
                                            )
#st_balanced_sum #this would not be expected to match Blanco because we used different seeds
#summarizing the data from the random samples
names <- c("Gold Standard (LT)", "ST Balanced")

kable(cbind(names , rbind(lt_gold_sum, st_balanced_sum)), digits = 2,
      caption = "Table 1.0: Balanced, Long-term & Short-term Sampling Campaign Summaries") %>% kable_styling()

```

```{r gold.standard.variability, warning = FALSE, echo = F}
#graphing the distribution of annual averages


balanced<- ggplot(st_balanced, aes(x = native_id, y = log_avg)) +
  geom_boxplot()+
  ggtitle("Figure 1:Short Term Balanced Sensor Log-transformed Annual Averages \n30 sets of 28 samples per sensor") +
  ylab("ln(NOx) ln(ppb)")

gold<- ggplot(long_term_gold, aes(x = native_id, y = log_avg)) +
  geom_point()+
  ggtitle("Gold Standard Sensor Log-transformed Annual Averages \n1 sample per sensor") +
  ylab("ln(NOx) ln(ppb)")


grid.arrange(balanced, gold)

grid.arrange(summer_state_var, fall_state_var, winter_state_var, spring_state_var)



```
These plots above show the variability in the data from all of the statewide data sets: gold standard, short term balanced, summer, fall, winter, and spring data sets. 

#Break up the histograms by season, look at NOx distribution by seasons


```{r, echo = F, include = T}

#Kinda crazy how much variability there is using unrestricted random samples
#We can choose which type of plot we like best - AG

# Abbie commented these histograms and used ggplot to re-make them so aesthetics could be added

#with(st_balanced, hist(annual_avg, breaks = 30, col = "blue", xlab = "NOx ppm" ))  
#with(st_balanced, hist(log_avg, breaks = 30, col = "seagreen", xlab = "log(NOx ppm)" ))

#with(long_term_gold, hist(annual_avg, breaks = 20, col = "blue", xlab = "NOx ppm" ))
#with(long_term_gold, hist(log_avg, breaks = 20, col = "seagreen", xlab = "log(NOx ppm)" ))

native <- long_term_gold %>%
ggplot(aes(annual_avg)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "lightblue", fill = "lightblue", alpha = 0.8) +
  geom_density(color = "purple") +
  labs(title = " Figure 2.0: Distribution of Annual Average NOx Concentrations",
       x = "Concentration on native scale (ppb)",
       y = "density"
       )

log <- long_term_gold %>%
ggplot(aes(log_avg)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "lightblue", fill = "lightblue", alpha = 0.8) +
  geom_density(color = "purple") +
  labs(title = "",
       x = "Concentration on natural log (ppb) scale",
       y = "density"
       )

native + log
#makes much more sense to work with the data on the log scale

#Should this whole chunk go to the appendix?

```
These histograms support our conclusion to log-transform our data before modeling with it. We see a rather obvious right tail on our native-scale NOx data, in what appears to be a log-normal distribution. The distribution becomes much more normal after log transformation (Right).

#obs vs preds for seasonal statewide models
```{r basic obs vs preds plot, echo = FALSE}
ggplot( st_balanced_CV, aes(x = log_avg, cvpreds)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("observations vs predictions - short term balanced campaign") +
  xlab("Observed log(NOx) annual averages") +
  ylab("Predicted log(NOx) annual averages ") +
  geom_abline(slope = 1.0) 

ggplot( st_summer_CV, aes(x = log_avg, cvpreds)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("observations vs predictions - short term Summer campaign") +
  xlab("Observed log(NOx) seasonal averages") +
  ylab("Predicted log(NOx) seasonal averages ") +
  geom_abline(slope = 1.0)

ggplot( st_fall_CV, aes(x = log_avg, cvpreds)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("observations vs predictions - short term Fall campaign") +
  xlab("Observed log(NOx) seasonal averages") +
  ylab("Predicted log(NOx) seasonal averages ") +
  geom_abline(slope = 1.0)

ggplot( st_winter_CV, aes(x = log_avg, cvpreds)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("observations vs predictions - short term Winter campaign") +
  xlab("Observed log(NOx) seasonal averages") +
  ylab("Predicted log(NOx) seasonal averages ") +
  geom_abline(slope = 1.0)

ggplot( st_spring_CV, aes(x = log_avg, cvpreds)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("observations vs predictions - short term Spring campaign") +
  xlab("Observed log(NOx) seasonal averages") +
  ylab("Predicted log(NOx) seasonal averages ") +
  geom_abline(slope = 1.0)

```


```{r, variability across LA county seasonal data sets}

LA_title <- "Variability in LA County Short Term Seasonal Data sets"
grid.arrange(top = LA_title, grobs = list(summer_LA, fall_LA, winter_LA, spring_LA, balanced_LA ))
#move to Appendix
```




```{r, warning = FALSE, echo = F, message = FALSE}

#summary figures - statewide

Title <- "Statewide Short Term Campaign Performances and Comparison with Gold Standard"
Methods <- " Method A = Campaign Predictions Compared to Gold Standard Obs\n
       Method B = Campaign Predictions Compared to its Observations\n
       Square points shows the Gold Standard's performance"

scatter_legend <- "x = Gold Standard predicted ln(NOx) (ln(ppb))
         \ny = Short Term Campaign Predicted ln(NOx) (ln(ppb))"

grid.arrange(top = Title, bottom = Methods, grobs= list(st_spring_boxplot, st_summer_boxplot, st_fall_boxplot, st_winter_boxplot, st_balanced_boxplot),
             ncol = 3, nrow = 3,
             layout_matrix = rbind(c(1,2,3),
                                   c(4, 5, NA)))

grid.arrange(top = Title, bottom = scatter_legend, grobs = list(st_spring_scatter, st_summer_scatter, st_fall_scatter, st_winter_scatter,
                                                                st_balanced_scatter), ncol = 3, nrow = 3,
             layout_matrix = rbind(c(1,2,3),
                                   c(4, 5, NA)))

#yikes! :)

#Abbie figured out grid.arrange, and made these much nicer plots.


```



```{r, warning = F, message = F, echo = F, include = F}
#summary figures - LA County

Title <- "LA County Short Term Campaign Performances and Comparison with Gold Standard"
Methods <- " Method A = Campaign Predictions Compared to Gold Standard Obs\n
       Method B = Campaign Predictions Compared to its Observations\n
       Square points shows the Gold Standard's performance"

scatter_legend <- "x = Gold Standard predicted ln(NOx) (ln(ppb))
         \ny = Short Term Campaign Predicted ln(NOx) (ln(ppb))"

grid.arrange(top = Title, bottom = Methods, grobs= list(st_spring_boxplot, st_summer_boxplot, st_fall_boxplot, st_winter_boxplot, st_balanced_boxplot),
             ncol = 3, nrow = 3,
             layout_matrix = rbind(c(1,2,3),
                                   c(4, 5, NA)))

grid.arrange(top = Title, bottom = scatter_legend, grobs = list(st_spring_scatter, st_summer_scatter, st_fall_scatter, st_winter_scatter,
                                                                st_balanced_scatter), ncol = 3, nrow = 3,
             layout_matrix = rbind(c(1,2,3),
                                   c(4, 5, NA)))

```


